---
title: "Visualizing the Uninsured: A Data Science Perspective on U.S. Health Coverage"
author: |
  Eduardo Sáenz-Messía Laborda  
  Catherina Mikhail (2847118)  
  Charlotte Craenen (2865572)  
  Hugo van As  
  Indy Pleijter  
  Julian Zeguers  
  Wies Polderman  
  Tutorial group 5, group 4 – J.F. Fitzgerald
date: "2025-06-24"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
options(repos = c(CRAN = "https://cran.rstudio.com/"))

packages <- c("rmarkdown", "tinytex", "rnaturalearth", "rnaturalearthdata", "maps", "tidyverse", "tidyr", "dplyr", "renv")

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) {
  install.packages(packages[!installed], repos = "https://cran.rstudio.com/")
}

invisible(lapply(packages, library, character.only = TRUE))

if (!tinytex::is_tinytex()) { 
  tinytex::install_tinytex()
}
```

Part 1 – Identify a Social Problem

## 1.1 Describe the Social Problem

The large number of uninsured people in the United States remains a deeply rooted social issue with serious individual and systemic consequences. As of 2022, over 27 million Americans—about one in every twelve—lacked health insurance (KFF, 2023). For those without coverage, the consequences can be severe. Uninsured individuals are far more likely to delay or skip necessary medical care due to cost, resulting in worsened health outcomes, preventable hospitalizations, and even premature death. Beyond the physical toll, the financial burden of paying for care out-of-pocket can be devastating, often leading to medical debt, credit problems, and long-term economic instability (Davis, 2007).

The problem doesn’t stop at the individual level. When millions are uninsured, the ripple effects are felt throughout the healthcare system. Hospitals and clinics often absorb the costs of uncompensated care, placing financial strain on public health resources and increasing costs for those who are insured. In the words of Davis (2007), “The uninsured are not just other people—they are our neighbors, co-workers, and family members,” and the consequences of their lack of coverage ripple through communities in both visible and invisible ways. The issue of uninsurance in America is not just a matter of personal misfortune; it reflects broader structural gaps in the nation’s approach to healthcare access and economic inclusion.

## 1.2 Provide Background on the Problem

Although the Affordable Care Act (ACA), passed in 2010, marked a major turning point by expanding access to coverage through Medicaid and subsidized insurance marketplaces, millions of Americans remain uninsured. According to the Kaiser Family Foundation, the majority of the uninsured today are low-income individuals, disproportionately people of color, and residents of states that have not expanded Medicaid (KFF, 2023). This uneven policy landscape has created deep geographic disparities in access to health insurance, especially across the South and parts of the Midwest.

One of the key structural challenges is that health insurance in the United States is closely tied to employment. As a result, people working in part-time, temporary, or informal jobs—who are often already economically vulnerable—are less likely to receive employer-sponsored insurance (KFF, 2023). This makes them particularly susceptible to losing coverage during periods of economic instability or job transitions. As Davis (2007) points out, many of the uninsured “fall through the cracks,” caught between eligibility thresholds and the high cost of private insurance. The problem is not only about whether coverage exists, but about whether it is affordable and accessible to the people who need it most.

In short, while progress has been made since the ACA, significant gaps in the system persist—especially for those at the intersection of poverty, racial inequity, and policy exclusion. These gaps leave millions without the basic security of knowing they can access healthcare when they need it.

# Part 2 – Describe and Acquire Data

## 2.1 Describe the Data sets

Our analysis is based on two main data sets: one from the U.S. Census Bureau and one from the Bureau of Economic Analysis (BEA), which together allowed us to explore patterns in health insurance coverage alongside economic conditions across U.S. states.

The primary data set comes from the U.S. Census Bureau’s *American Community Survey (ACS) 1-Year Estimates*, specifically the table *Selected Characteristics of the Uninsured in the United States*. We used data from 2010 to 2023, excluding 2020 due to a lack of 1-year estimates that year. We also excluded eight small states and territories—Delaware, the District of Columbia, Hawaii, Puerto Rico, North Dakota, Rhode Island, Vermont, and Wyoming—because they are not consistently included in the 1-year estimates and are instead only available in the 5-year version. Since our goal was to analyze annual trends, we relied on the 1-year estimates for their timeliness and ability to reflect short-term changes, such as those driven by economic shifts or policy reforms.

This ACS data set provides state-level figures on the civilian non-institutionalized population, including both the total population and the number of uninsured individuals. It also includes a wide range of demographic and socioeconomic characteristics. For our analysis, we focused only on the total population and household income (inflation-adjusted), which were essential for calculating the uninsured share and understanding economic context.

To supplement this, we used economic data from the BEA, specifically the table *SAGDP1 – State Annual Gross Domestic Product Summary*. This data set provides real GDP figures in millions of chained 2017 dollars for each state and aggregate regions, covering the years 1997 through 2024. Using inflation-adjusted GDP allows for consistent comparisons over time and across states.

Together, these two data sets gave us a comprehensive view of how health insurance coverage has evolved across the U.S., and how it may be linked to broader economic trends at the state level.

## 2.2 Import and Prepare the Dataset

**First data set:** **"Selected Characteristics of The Uninsured in the U.S."**

-   Here we imported the first data set from the U.S. Census Bureau and named them "dataset_uninsured_population\_*year"* from 2010-2023, excl 2020.

```{r import-data, include=FALSE}
# The first dataset is from the dataset "selected characteristics of the uninsured in the US from 2010-2023, exclduing 2020 from the US Census Bureau. These are the ACS one-year estimates.

#2010
ACSST1Y2010_S2702_2025_06_03T123607 <- read_csv("data insured population 2010-2023/ACSST1Y2010.S2702-2025-06-03T123607.csv")
dataset_uninsured_population_2010 = read_csv("data insured population 2010-2023/ACSST1Y2010.S2702-2025-06-03T123607.csv")

#2011
ACSST1Y2011_S2702_2025_06_03T123539 <- read_csv("data insured population 2010-2023/ACSST1Y2011.S2702-2025-06-03T123539.csv")
dataset_uninsured_population_2011 = read_csv("data insured population 2010-2023/ACSST1Y2011.S2702-2025-06-03T123539.csv")

#2012
ACSST1Y2012_S2702_2025_06_03T123441 <- read_csv("data insured population 2010-2023/ACSST1Y2012.S2702-2025-06-03T123441.csv")
dataset_uninsured_population_2012 = read_csv("data insured population 2010-2023/ACSST1Y2012.S2702-2025-06-03T123441.csv")

#2013
ACSST1Y2013_S2702_2025_06_03T123404 <- read_csv("data insured population 2010-2023/ACSST1Y2013.S2702-2025-06-03T123404.csv")
dataset_uninsured_population_2013 = read_csv("data insured population 2010-2023/ACSST1Y2013.S2702-2025-06-03T123404.csv")

#2014
ACSST1Y2014_S2702_2025_06_03T123336 <- read_csv("data insured population 2010-2023/ACSST1Y2014.S2702-2025-06-03T123336.csv")
dataset_uninsured_population_2014 = read_csv("data insured population 2010-2023/ACSST1Y2014.S2702-2025-06-03T123336.csv")

#2015
ACSST1Y2015_S2702_2025_06_03T123312 <- read_csv("data insured population 2010-2023/ACSST1Y2015.S2702-2025-06-03T123312.csv")
dataset_uninsured_population_2015 = read_csv("data insured population 2010-2023/ACSST1Y2015.S2702-2025-06-03T123312.csv")

#2016
ACSST1Y2016_S2702_2025_06_03T123232 <- read_csv("data insured population 2010-2023/ACSST1Y2016.S2702-2025-06-03T123232.csv")
dataset_uninsured_population_2016 = read_csv("data insured population 2010-2023/ACSST1Y2016.S2702-2025-06-03T123232.csv")

#2017
ACSST1Y2017_S2702_2025_06_03T123200 <- read_csv("data insured population 2010-2023/ACSST1Y2017.S2702-2025-06-03T123200.csv")
dataset_uninsured_population_2017 = read_csv("data insured population 2010-2023/ACSST1Y2017.S2702-2025-06-03T123200.csv")

#2018
ACSST1Y2018_S2702_2025_06_03T123128 <- read_csv("data insured population 2010-2023/ACSST1Y2018.S2702-2025-06-03T123128.csv")
dataset_uninsured_population_2018 = read_csv("data insured population 2010-2023/ACSST1Y2018.S2702-2025-06-03T123128.csv")

#2019
ACSST1Y2019_S2702_2025_06_03T123018 <- read_csv("data insured population 2010-2023/ACSST1Y2019.S2702-2025-06-03T123018.csv")
dataset_uninsured_population_2019 = read_csv("data insured population 2010-2023/ACSST1Y2019.S2702-2025-06-03T123018.csv")

#2021
ACSST1Y2021_S2702_2025_06_03T122952 <- read_csv("data insured population 2010-2023/ACSST1Y2021.S2702-2025-06-03T122952.csv")
dataset_uninsured_population_2021 = read_csv("data insured population 2010-2023/ACSST1Y2021.S2702-2025-06-03T122952.csv")

#2022
ACSST1Y2022_S2702_2025_06_03T122917 <- read_csv("data insured population 2010-2023/ACSST1Y2022.S2702-2025-06-03T122917.csv")
dataset_uninsured_population_2022 = read_csv("data insured population 2010-2023/ACSST1Y2022.S2702-2025-06-03T122917.csv")

#2023
ACSST1Y2023_S2702_2025_06_04T165512 <- read_csv("data insured population 2010-2023/ACSST1Y2023.S2702-2025-06-04T165512.csv")
dataset_uninsured_population_2023 = read_csv("data insured population 2010-2023/ACSST1Y2023.S2702-2025-06-04T165512.csv")

```

**Second Data set: "SAGDP1 State Annual Gross Domestic Product Summary", Statistic: "Real GDP (millions of chained 2017 dollars)"**

-   Here we imported the second data set from the Bureau of Economic Analysis of the annual GDP of the states from 2009-2023 and named it "table_GDP"

```{r import second dataset, figure.show="asis", include=F}
#The second dataset is from the Bureau of Economic Analysis and is the annual real GDP of the 44 states we include in our analysis from the year 2009-2023, excluding 2020
annual_gdp_Table_4_ <- read_csv("data GDP/annual gdp - Table (4).csv")

#renaming the table
table_GDP = annual_gdp_Table_4_
```

## Data cleaning: Data sets of Uninsured Population and The Creation of The New Variable: The Uninsured Share

To prepare the data for analysis, we began by removing all margins of error from the original Census data sets. While margins of error are important for assessing statistical precision, they were not necessary for our purposes, which focused on comparing overall trends in health insurance coverage across states and years. Removing them helped simplify the data set and reduce complexity without compromising our analytical goals.

A more structural challenge involved the inclusion of smaller states and territories. Since 2016, the Census Bureau’s 1-year estimates no longer cover certain low-population areas, meaning they are missing from some years in our timespan. To ensure consistency across time, we decided to exclude eight such places entirely from our analysis. These were Delaware, the District of Columbia, Hawaii, Puerto Rico, North Dakota, Rhode Island, Vermont, and Wyoming. By removing them, we ensured that our data set included only states that were present in every applicable year from 2010 to 2023 (with the exception of 2020, when no data was published due to the COVID-19 pandemic).

After resolving issues of consistency and coverage, we turned our attention to the structure of the data itself. Each annual data set included a wide range of metadata—demographic and socioeconomic characteristics that, while valuable for other types of research, were not essential to our primary analysis of insurance coverage trends. We removed variables related to age, sex, race and Hispanic or Latino origin, nativity and U.S. citizenship status, disability status, residence one year ago, educational attainment, employment status, work experience, civilian non-institutionalized workers aged 16 and over, earnings in the past 12 months, and the ratio of income to the poverty level in the past 12 months. From all this, we retained only the total civilian non-institutionalized population and household income (adjusted for inflation), which were directly relevant to our questions about insurance and economic context.

To ensure the data was ready for calculations, we then converted all numeric values that had been stored as character strings—often due to formatting like commas—into actual numeric variables. This was a necessary step for performing any reliable mathematical operations.

One of the most important transformations in our cleaning process was the creation of a new variable: uninsured share. The original data provided the total population and the total number of uninsured individuals for each state and year. To make this more interpretable and comparable, we calculated the uninsured share by dividing the number of uninsured people by the total population and multiplying the result by 100. This gave us a percentage that reflects the proportion of each state’s population without health insurance. Expressing the data this way allows for easier comparisons across time and across states, regardless of differences in population size. It helps illuminate broader patterns in insurance coverage that would be difficult to spot using absolute numbers alone.

After all these cleaning and transformation steps, we combined the individual yearly data sets into a single, comprehensive data set covering the period from 2010 to 2023 (excluding 2020). This final merged file allowed us to analyze long-term trends and relationships in a consistent and structured way.

```{r datacleaning of the yearly datasets Uninsured Population 2010,2023, include=FALSE}
# This is me cleaning out the data insured population from 2010-2016 getting out the margin of error
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)

###############################################################################

#This is me cleaning out the columns of Delaware
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)

#This is me cleaning out the columns of District of Columbia
dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)

#This is me cleaning out the columns of Hawaii
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of North Dakota
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Puerto Rico
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Rhode Island
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Vermont
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Wyoming
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)

###############################################################################

#Now we are cleaning out the unnecessary metadata in the rows
#this is me cleaning out the rows with unnecessary metadata and keeping household income levels and the total population numbers
dataset_uninsured_population_2010 <- dataset_uninsured_population_2010[-c(2:104),]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 <- dataset_uninsured_population_2011[-c(2:104),]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 <- dataset_uninsured_population_2012[-c(2:104),]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 <- dataset_uninsured_population_2013[-c(2:104),]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 <- dataset_uninsured_population_2014[-c(2:104),]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 <- dataset_uninsured_population_2015[-c(2:104),]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 <- dataset_uninsured_population_2016[-c(2:104),]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 <- dataset_uninsured_population_2017[-c(2:104),]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 <- dataset_uninsured_population_2018[-c(2:104),]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 <- dataset_uninsured_population_2019[-c(2:104),]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 <- dataset_uninsured_population_2021[-c(2:104),]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 <- dataset_uninsured_population_2022[-c(2:104),]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 <- dataset_uninsured_population_2023[-c(2:104),]
View(dataset_uninsured_population_2023)

dataset_uninsured_population_2010 <- dataset_uninsured_population_2010[-c(10:17),]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 <- dataset_uninsured_population_2011[-c(10:17),]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 <- dataset_uninsured_population_2012[-c(10:17),]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 <- dataset_uninsured_population_2013[-c(10:17),]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 <- dataset_uninsured_population_2014[-c(10:17),]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 <- dataset_uninsured_population_2015[-c(10:17),]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 <- dataset_uninsured_population_2016[-c(10:17),]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 <- dataset_uninsured_population_2017[-c(10:17),]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 <- dataset_uninsured_population_2018[-c(10:17),]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 <- dataset_uninsured_population_2019[-c(10:17),]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 <- dataset_uninsured_population_2021[-c(10:17),]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 <- dataset_uninsured_population_2022[-c(10:17),]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 <- dataset_uninsured_population_2023[-c(10:17),]
View(dataset_uninsured_population_2023)

################################################################################

#As the values in these yearly datasets are stored as characters, we are now going to coerce them into numeric values so that we can create a new variable afterwards.
#The values in the datasets of the uninsured population are stored as characters. Here I am going to get rid of "" so I can store them as numeric.

# Keep the first column as-is
first_col <- dataset_uninsured_population_2010[[1]]

# Clean and convert the rest of the columns
# Keep the first column as-is
first_col <- dataset_uninsured_population_2010[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2010_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2010[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2010_cleaned <- cbind(first_col, dataset_uninsured_population_2010_numeric)

#rename the first column
colnames(dataset_uninsured_population_2010_cleaned)[1] <- colnames(dataset_uninsured_population_2010)[1]

View(dataset_uninsured_population_2010_cleaned)
class(dataset_uninsured_population_2010_cleaned)
class(dataset_uninsured_population_2010_cleaned[[2]])




# Keep the first column as-is
first_col <- dataset_uninsured_population_2011[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2011_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2011[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2011_cleaned <- cbind(first_col, dataset_uninsured_population_2011_numeric)

#rename the first column
colnames(dataset_uninsured_population_2011_cleaned)[1] <- colnames(dataset_uninsured_population_2011)[1]

View(dataset_uninsured_population_2011_cleaned)
class(dataset_uninsured_population_2011_cleaned)
class(dataset_uninsured_population_2011_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2012[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2012_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2012[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2012_cleaned <- cbind(first_col, dataset_uninsured_population_2012_numeric)

#rename the first column
colnames(dataset_uninsured_population_2012_cleaned)[1] <- colnames(dataset_uninsured_population_2012)[1]

View(dataset_uninsured_population_2012_cleaned)
class(dataset_uninsured_population_2012_cleaned)
class(dataset_uninsured_population_2012_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2013[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2013_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2013[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2013_cleaned <- cbind(first_col, dataset_uninsured_population_2013_numeric)

#rename the first column
colnames(dataset_uninsured_population_2013_cleaned)[1] <- colnames(dataset_uninsured_population_2013)[1]

View(dataset_uninsured_population_2013_cleaned)
class(dataset_uninsured_population_2013_cleaned)
class(dataset_uninsured_population_2013_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2014[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2014_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2014[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2014_cleaned <- cbind(first_col, dataset_uninsured_population_2014_numeric)

#rename the first column
colnames(dataset_uninsured_population_2014_cleaned)[1] <- colnames(dataset_uninsured_population_2014)[1]

View(dataset_uninsured_population_2014_cleaned)
class(dataset_uninsured_population_2014_cleaned)
class(dataset_uninsured_population_2014_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2015[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2015_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2015[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2015_cleaned <- cbind(first_col, dataset_uninsured_population_2015_numeric)

#rename the first column
colnames(dataset_uninsured_population_2015_cleaned)[1] <- colnames(dataset_uninsured_population_2015)[1]

View(dataset_uninsured_population_2015_cleaned)
class(dataset_uninsured_population_2015_cleaned)
class(dataset_uninsured_population_2015_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2016[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2016_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2016[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2016_cleaned <- cbind(first_col, dataset_uninsured_population_2016_numeric)

#rename the first column
colnames(dataset_uninsured_population_2016_cleaned)[1] <- colnames(dataset_uninsured_population_2016)[1]

View(dataset_uninsured_population_2016_cleaned)
class(dataset_uninsured_population_2016_cleaned)
class(dataset_uninsured_population_2016_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2017[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2017_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2017[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2017_cleaned <- cbind(first_col, dataset_uninsured_population_2017_numeric)

#rename the first column
colnames(dataset_uninsured_population_2017_cleaned)[1] <- colnames(dataset_uninsured_population_2017)[1]

View(dataset_uninsured_population_2017_cleaned)
class(dataset_uninsured_population_2017_cleaned)
class(dataset_uninsured_population_2017_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2018[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2018_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2018[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2018_cleaned <- cbind(first_col, dataset_uninsured_population_2018_numeric)

#rename the first column
colnames(dataset_uninsured_population_2018_cleaned)[1] <- colnames(dataset_uninsured_population_2018)[1]

View(dataset_uninsured_population_2018_cleaned)
class(dataset_uninsured_population_2018_cleaned)
class(dataset_uninsured_population_2018_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2019[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2019_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2019[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2019_cleaned <- cbind(first_col, dataset_uninsured_population_2019_numeric)

#rename the first column
colnames(dataset_uninsured_population_2019_cleaned)[1] <- colnames(dataset_uninsured_population_2019)[1]

View(dataset_uninsured_population_2019_cleaned)
class(dataset_uninsured_population_2019_cleaned)
class(dataset_uninsured_population_2019_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2021[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2021_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2021[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2021_cleaned <- cbind(first_col, dataset_uninsured_population_2021_numeric)

#rename the first column
colnames(dataset_uninsured_population_2021_cleaned)[1] <- colnames(dataset_uninsured_population_2021)[1]

View(dataset_uninsured_population_2021_cleaned)
class(dataset_uninsured_population_2021_cleaned)
class(dataset_uninsured_population_2021_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2022[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2022_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2022[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2022_cleaned <- cbind(first_col, dataset_uninsured_population_2022_numeric)

#rename the first column
colnames(dataset_uninsured_population_2022_cleaned)[1] <- colnames(dataset_uninsured_population_2022)[1]

View(dataset_uninsured_population_2022_cleaned)
class(dataset_uninsured_population_2022_cleaned)
class(dataset_uninsured_population_2022_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2023[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2023_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2023[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2023_cleaned <- cbind(first_col, dataset_uninsured_population_2023_numeric)

#rename the first column
colnames(dataset_uninsured_population_2023_cleaned)[1] <- colnames(dataset_uninsured_population_2023)[1]

View(dataset_uninsured_population_2023_cleaned)
class(dataset_uninsured_population_2023_cleaned)
class(dataset_uninsured_population_2023_cleaned[[2]])

################################################################################

#Now that the values in the yearly datasets are stored as numeric values but the first column as the first column contains the names of the states, we can now create are first new variable, which is the uninsured share:

#Here I am trying to compute and create a new variable, which is the share of uninsured population per state and per year

# Copy original data 2010
df_pop_2010 <- dataset_uninsured_population_2010_cleaned
col_names <- colnames(df_pop_2010)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2010[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2010[[total_col]][1]) && df_pop_2010[[total_col]][1] != 0) {
      df_pop_2010[[share_col]][1] <- df_pop_2010[[uninsured_col]][1] / df_pop_2010[[total_col]][1]
    } else {
      df_pop_2010[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2010 <- df_pop_2010[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2010), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2010[1, share_cols])

View(df_pop_2010)

#################################################################################

# Copy original data 2011
df_pop_2011 <- dataset_uninsured_population_2011_cleaned
col_names <- colnames(df_pop_2011)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2011[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2011[[total_col]][1]) && df_pop_2011[[total_col]][1] != 0) {
      df_pop_2011[[share_col]][1] <- df_pop_2011[[uninsured_col]][1] / df_pop_2011[[total_col]][1]
    } else {
      df_pop_2011[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2011 <- df_pop_2011[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2011), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2011[1, share_cols])

View(df_pop_2011)


###############################################################################


# Copy original data 2012
df_pop_2012 <- dataset_uninsured_population_2012_cleaned
col_names <- colnames(df_pop_2012)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2012[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2012[[total_col]][1]) && df_pop_2012[[total_col]][1] != 0) {
      df_pop_2012[[share_col]][1] <- df_pop_2012[[uninsured_col]][1] / df_pop_2012[[total_col]][1]
    } else {
      df_pop_2012[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2012 <- df_pop_2012[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2012), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2012[1, share_cols])


###############################################################################


# Copy original data 2013
df_pop_2013 <- dataset_uninsured_population_2013_cleaned
col_names <- colnames(df_pop_2013)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2013[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2013[[total_col]][1]) && df_pop_2013[[total_col]][1] != 0) {
      df_pop_2013[[share_col]][1] <- df_pop_2013[[uninsured_col]][1] / df_pop_2013[[total_col]][1]
    } else {
      df_pop_2013[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2013 <- df_pop_2013[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2013), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2013[1, share_cols])


################################################################################


# Copy original data 2014
df_pop_2014 <- dataset_uninsured_population_2014_cleaned
col_names <- colnames(df_pop_2014)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2014[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2014[[total_col]][1]) && df_pop_2014[[total_col]][1] != 0) {
      df_pop_2014[[share_col]][1] <- df_pop_2014[[uninsured_col]][1] / df_pop_2014[[total_col]][1]
    } else {
      df_pop_2014[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2014 <- df_pop_2014[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2014), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2014[1, share_cols])


################################################################################


# Copy original data 2015
df_pop_2015 <- dataset_uninsured_population_2015_cleaned
col_names <- colnames(df_pop_2015)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2015[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2015[[total_col]][1]) && df_pop_2015[[total_col]][1] != 0) {
      df_pop_2015[[share_col]][1] <- df_pop_2015[[uninsured_col]][1] / df_pop_2015[[total_col]][1]
    } else {
      df_pop_2015[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2015 <- df_pop_2015[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2015), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2015[1, share_cols])


###############################################################################


# Copy original data 2016
df_pop_2016 <- dataset_uninsured_population_2016_cleaned
col_names <- colnames(df_pop_2016)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2016[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2016[[total_col]][1]) && df_pop_2016[[total_col]][1] != 0) {
      df_pop_2016[[share_col]][1] <- df_pop_2016[[uninsured_col]][1] / df_pop_2016[[total_col]][1]
    } else {
      df_pop_2016[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2016 <- df_pop_2016[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2016), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2016[1, share_cols])

################################################################################

#something went wrong here with 2017 and 2018, so the function that worked is much more below as the column names in the years of 2017-2018 were named differently than the years before

# Copy original data 2017
df_pop_2017 <- dataset_uninsured_population_2017_cleaned
col_names <- colnames(df_pop_2017)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2017[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2017[[total_col]][1]) && df_pop_2017[[total_col]][1] != 0) {
      df_pop_2017[[share_col]][1] <- df_pop_2017[[uninsured_col]][1] / df_pop_2017[[total_col]][1]
    } else {
      df_pop_2017[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2017 <- df_pop_2017[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2017), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2017[1, share_cols])

# Copy original data 2018
df_pop_2018<- dataset_uninsured_population_2018_cleaned
col_names <- colnames(df_pop_2018)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2018[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2018[[total_col]][1]) && df_pop_2018[[total_col]][1] != 0) {
      df_pop_2018[[share_col]][1] <- df_pop_2018[[uninsured_col]][1] / df_pop_2018[[total_col]][1]
    } else {
      df_pop_2018[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2018 <- df_pop_2018[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2018), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2018[1, share_cols])

View(df_pop_2017)
View(df_pop_2016)
View(df_pop_2018)

#I have seen that it did not work for the years of 2017 and 2018 so I have to see what the names of the columns are
grep("Alabama", colnames(dataset_uninsured_population_2017_cleaned), value = TRUE)
grep("Alabama", colnames(dataset_uninsured_population_2018_cleaned), value = TRUE)

#Here I am starting again from 2017

# Make a working copy of your data 2017
df_pop_2017 <- dataset_uninsured_population_2017_cleaned
col_names <- colnames(df_pop_2017)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2017[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2017[[total_col]][1]
    uninsured_val <- df_pop_2017[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2017[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2017 <- df_pop_2017[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2017), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2017[1, share_cols])

#now it worked and I can work further on 2018-2023 except for 2020 as we do not have the data of that

# Make a working copy of your data 2018
df_pop_2018 <- dataset_uninsured_population_2018_cleaned
col_names <- colnames(df_pop_2018)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2018[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2018[[total_col]][1]
    uninsured_val <- df_pop_2018[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2018[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2018 <- df_pop_2018[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2018), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2018[1, share_cols])


################################################################################


# Make a working copy of your data 2019
df_pop_2019 <- dataset_uninsured_population_2019_cleaned
col_names <- colnames(df_pop_2019)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2019[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2019[[total_col]][1]
    uninsured_val <- df_pop_2019[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2019[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2019 <- df_pop_2019[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2019), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2019[1, share_cols])


################################################################################


# Make a working copy of your data 2021
df_pop_2021 <- dataset_uninsured_population_2021_cleaned
col_names <- colnames(df_pop_2021)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2021[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2021[[total_col]][1]
    uninsured_val <- df_pop_2021[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2021[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2021 <- df_pop_2021[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2021), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2021[1, share_cols])

###############################################################################


# Make a working copy of your data 2022
df_pop_2022 <- dataset_uninsured_population_2022_cleaned
col_names <- colnames(df_pop_2022)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2022[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2022[[total_col]][1]
    uninsured_val <- df_pop_2022[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2022[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2022 <- df_pop_2022[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2022), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2022[1, share_cols])


################################################################################

# Make a working copy of your data 2023
df_pop_2023 <- dataset_uninsured_population_2023_cleaned
col_names <- colnames(df_pop_2023)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2023[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2023[[total_col]][1]
    uninsured_val <- df_pop_2023[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2023[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2023 <- df_pop_2023[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2023), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2023[1, share_cols])

################################################################################

#Now we have the new variable in the seperate datasets and now we are going to merge these yearly datasets into 1 dataset

# Define years
years_early <- 2010:2016
years_late <- c(2017:2019, 2021:2023)

# Function to process one year of data
process_df <- function(df, year, is_early) {
  label_col <- names(df)[1]
  state_cols <- names(df)[!names(df) %in% c(label_col, "Unnamed: 0")]
  states <- unique(str_extract(state_cols, "^[^.]+"))
  
  # Set correct suffixes depending on year
  if (is_early) {
    pop_suffix <- "..Total.Civilian.Noninstitutionalized.Population..Estimate"
    unins_suffix <- "..Uninsured.Population..Estimate"
    share_suffix <- "..Uninsured.Share"
  } else {
    pop_suffix <- "..Total..Estimate"
    unins_suffix <- "..Total.Uninsured..Estimate"
    share_suffix <- "..Uninsured.Share"
  }
  
  # Extract for each state
  rows <- lapply(states, function(state) {
    pop_col <- paste0(state, pop_suffix)
    unins_col <- paste0(state, unins_suffix)
    share_col <- paste0(state, share_suffix)
    
    # Get only total population row
    total_row <- df %>%
      filter(str_trim(.data[[label_col]]) == "Total population")
    
    if (nrow(total_row) == 0 || !(pop_col %in% names(df))) return(NULL)
    
    tibble(
      state = state,
      year = year,
      total_population = as.numeric(total_row[[pop_col]]),
      total_uninsured = as.numeric(total_row[[unins_col]]),
      uninsured_share = as.numeric(total_row[[share_col]])
    )
  })
  
  bind_rows(rows)
}

# Process all early years
dfs_early <- mget(paste0("df_pop_", years_early))
panel_early <- map2_dfr(dfs_early, years_early, ~process_df(.x, .y, is_early = TRUE))

# Process all late years
dfs_late <- mget(paste0("df_pop_", years_late))
#panel_late <- map2_dfr(dfs_late, years_late, ~process_df(.x, .y, is_early = FALSE))

early_states <- bind_rows(
  dfs_early, .id = "year") %>%
  pivot_longer(
    -c(year, "Label (Grouping)"),
    names_to = "tmp",
    values_to = "value") %>%
  separate(tmp,
           into = c("state", "series", "measure"),
           sep = "\\.\\.",
           remove = TRUE)

late_states <- bind_rows(
  dfs_late, .id = "year") %>%
  pivot_longer(
    -c(year, "Label (Grouping)"),
    names_to = "tmp",
    values_to = "value") %>%
  separate(tmp,
           into = c("state", "series", "measure"),
           sep = "\\.\\.",
           remove = TRUE)
# View

all_states <- rbind(late_states, early_states)

head(all_states)
###############################################################################

#cleaning out df_pop from the column of years to only remain the year numbers

unique(all_states$year)

all_states$year <- str_remove(all_states$year, "df_pop_")

#Now I am coercing the string into numeric so plotting becomes easier

all_states$year <- as.numeric(all_states$year)
str(all_states)

################################################################################
#Here I am multiplying the uninsured share by 100, such that they are representing percentages
all_states_100 = all_states %>%
  mutate(value = 
           ifelse(grepl("Uninsured.Share",
                        series), value*100, value))

#This is the dataset for the Uninsured Population we can use for the visualizations
#renaming a column to make it easier "Label (Grouping)" to "income_group"

library(dplyr)
all_states_100 = all_states_100 %>%
  rename(income_group = `Label (Grouping)`)
```

## Data cleaning: Annual GDP Table and The Creation of The New Variable: Annual Real GDP Growth per year and state

For the economic data from the Bureau of Economic Analysis, we worked with the *SAGDP1 – State Annual Gross Domestic Product Summary* table, which reports real GDP per state in millions of chained 2017 dollars. Before analysis, we cleaned this data set to align it with the structure and timeframe of our health insurance data. First, we removed all columns corresponding to years outside our scope—specifically, 1997 through 2008, as well as 2020 and 2024. We also filtered out states and regions that were not included in our main data set or that represented aggregate regions rather than individual states. Specifically, we excluded Delaware, District of Columbia, Hawaii, Vermont, Wyoming, and Rhode Island, as well as the regional aggregates: New England, Mideast, Great Lakes, Plains, Southeast, Southwest, Rocky Mountain, and Far West. Also here, the values were stored as "double", so we coerced the values to numeric, so that we are able to compute our new variable. After these adjustments, we renamed the cleaned data set `table_GDP` for use in our analysis.

From this cleaned table, we created a new variable: annual real GDP growth, calculated for each state and year by measuring the year-over-year percentage change in real GDP. This variable allowed us to capture how quickly or slowly each state’s economy was growing over time. Including GDP growth in our analysis adds important context: it helps us explore whether changes in economic performance are linked to shifts in health insurance coverage. For example, we can examine whether periods of economic expansion correspond with improvements in coverage rates. By combining this with our uninsurance data, we gain a more nuanced understanding of how economic conditions may influence access to healthcare.

```{r include=FALSE}
#Here I am making a new variable which is the GDP growth rate of the states

#renaming the table
table_GDP = annual_gdp_Table_4_

#cleaning out the table of the years 1997:2008, 2020 and 2024 and removing the states/regions that are not included in our main data set

table_GDP = annual_gdp_Table_4_[, !(names(annual_gdp_Table_4_) %in% c("1997", "1998","1999","2000", "2001","2002","2003","2004","2005","2006","2007","2008","2020","2024"))]

table_GDP = table_GDP[-c(9,10,13,36,41,47,52,54,55,56,57,58,59,60),]
table_GDP = table_GDP[-c(1),]
table_GDP = table_GDP[-c(45),]

#class coercion of double to numeric of all columns but the first column
table_GDP[, -1] = lapply(table_GDP[, -1], as.numeric)

#checking the class
sapply(table_GDP, class)

#yes, all the numbers are now numeric instead of the first column with the names of the states


# Pivot longer
gdp_long <- table_GDP %>%
  pivot_longer(
    cols = -GeoName,
    names_to = "year",
    values_to = "gdp"
  ) %>%
  mutate(year = as.integer(year))

View(table_GDP)
View(gdp_long)

# Calculate GDP growth rate
gdp_growth <- gdp_long %>%
  group_by(GeoName) %>%
  arrange(year) %>%
  mutate(growth_rate = (gdp - lag(gdp)) / lag(gdp) * 100)

View(gdp_growth)

#delete rows from 2009
gdp_growth = gdp_growth[-c(1:45), ]
```

# Part 3 – Visualize and Analyze the Data

## 3.1 Create Initial Visualizations

**Spatial Variation Visualization, U.S. Share of The Uninsured Population in 2023**

-   describe here that we mapped here the spatial variation of the uninsured population of the states of which we have the ACS one year estimates of and that we first created a subset of 2023 of the big dataset all_states_100 and that we plotted Alaska seperately otherwise the whole map would be too small to visualize.

```{r visualize spatial, include=FALSE, fig.show = "asis"}
#Here I am making a sub dataframe for the spatial visualization of 2023 of all the states with available data

spatial_df = all_states %>%
  filter(year == 2023, series == "Uninsured.Share") %>%
  select(state, value)

#deleting unnecessary rows
spatial_df = spatial_df[-c(45:396),]

#Here I am trying to plot the spatial subset

#getting the US map
us_map = map_data("state")

#clean up states names in both datasets
us_map$region = tolower(us_map$region)
spatial_df$state = tolower(spatial_df$state)

#join the map data with the subset data
us_map = left_join(us_map, spatial_df, by = c("region" = "state"))

#plot the map
ggplot(us_map, aes(x = long, y = lat, group = group, fill = value)) + 
  geom_polygon (color = "black") +
  scale_fill_gradient(low = "lightblue", high = "red", na.value = "white")+
  ggtitle("Share of Uninsured Population in the US, 2023")+
  labs(fill = "(%) Share of Uninsured Population") + 
  theme_void() 

#I still need the state Alaska but the package maps does not include the state Alaska

# I am trying it here again including Alaska with another package

library(rnaturalearth)
library(ggplot2)
library(dplyr)

#get US shapefile
us_states = ne_states(country = "United States of America", returnclass = "sf")

#convert state names into lowercase letters such that they match
us_states$name = tolower(us_states$name)

alaska = us_states[us_states$name == "alaska",]

#merge data with shapefile
alaska_data = alaska %>%
  left_join(spatial_df, by = c("name" = "state"))

#plot the map with the data
ggplot(alaska_data) + 
  geom_sf(aes(fill = value)) + 
  theme_void() + 
  scale_fill_gradient( low = "lightblue", high = "red", na.value = "white")+
  theme(legend.position = "none")
```

**Temporal Variation Visualization, Texas's Share of Uninsured Population and Texas's GDP growth rate over time**

-   describe that Texas stood out in our spatial variation analysis and that we wanted to dig deeper into Texas's share of uninsured population over time from 2010 until 2023, except for 2020 and that we wanted to see if economic growth played a role in this.

```{r visualization temporal, include=FALSE, fig.show = "asis"}
#Here I am making a sub dataframe for the temporal visualization with the GDP growth rate of Texas and another subset of the share of uninsured population in the US over the years

#subset Texas share of uninsured population
texas_share = all_states_100 %>%
  filter(state == "Texas",
         year %in% 2010:2023,
         series == "Uninsured.Share")

#cleaning out the NA rows of subset Texas
texas_share = texas_share[-c(2:9),]
texas_share = texas_share[-c(3:10),]
texas_share = texas_share[-c(4:11),]
texas_share = texas_share[-c(5:12),]
texas_share = texas_share[-c(6:13),]
texas_share = texas_share[-c(7:14),]
texas_share = texas_share[-c(8:15),]
texas_share = texas_share[-c(9:16),]
texas_share = texas_share[-c(10:17),]
texas_share = texas_share[-c(11:18),]
texas_share = texas_share[-c(12:19),]
texas_share = texas_share[-c(13:20),]
texas_share = texas_share[-c(14:21),]


#subset GDP growth of texas of the years 2010-2023 except for 2020
library(dplyr)
texas_gdp = gdp_growth %>%
  filter(GeoName == "Texas",
         year %in% 2010:2023) %>%
  select(year, growth_rate)

#Here I will plot the temporal visualization of Texas over the years 2010-2023 except for year 2020

#the subsets I will use are
View(texas_share)
View(texas_gdp)

library(ggplot2)
library(dplyr)

#Add a type column to distinguish between GDP growth rate and Uninsured Share
texas_gdp$type = "GDP growth rate"
texas_share$type = "Uninsured Share"

#rename the columns to match
texas_share = texas_share %>%
  rename(gdp = value)

#combine the 2 dataframes
combined_data_texas = bind_rows(texas_gdp %>%
                                  select(year, gdp = growth_rate, type),
                                texas_share %>%
                                  select(year, gdp, type))

################################################################################

#plotting the lines from the 2 dataframes

ggplot() + 
  geom_line(data = texas_gdp, aes(x = year, y = growth_rate, color = "GDP Growth Rate"), size =1) +
  geom_line(data = texas_share, aes(x = year, y = gdp, color = "Uninsured Share"), size = 1) + 
  labs(title = "GDP growth rate and Uninsured Share in Texas 2010 - 2023, except 2020", x = "Year", y = "Percentage Points", color = "Type") +
  scale_y_continuous(breaks = seq(-2.5,25, by = 2.5), limits = c(-2.5,25)) + 
  scale_x_continuous(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2021,2022,2023)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

**Sub-Group Variation Visualization, The Uninsured Rates of the Middle-Income Group in 2017 in the U.S.**

-   describe that already the KFF (2024) found that the uninsured people in the US are more likely to be low income and that we were wondering what the state of the middle-income group is as they earn too much for Medicaid but not enough for private insurance. 2017 is the year after Trump became president in 2016.

```{r visualization subgroup, include=FALSE, fig.show = "asis"}
#create a subset of data 2017 for total unsinred share in the states

library(dplyr)

subgroup_total = all_states_100 %>%
  filter(year == 2017,
         series == "Uninsured.Share",
         income_group == "Total civilian noninstitutionalized population")
# create a subset data of 2017 and middle income
middle_income = all_states_100[ all_states_100$income_group == ("$25,000 to $49,999"),]
middle_income = middle_income[middle_income$series == "Total.Uninsured",]
#subset that will be used is middle_income of 2017
#calculate quartiles
quartiles = quantile(middle_income$value, probs = c(0, 0.25, 0.5, 0.75, 1))

middle_income_1 = middle_income[middle_income$value >= quartiles[1] & middle_income$value <= quartiles[2],]
middle_income_2 = middle_income[middle_income$value > quartiles[2] & middle_income$value <= quartiles [3],]
middle_income_3 = middle_income[middle_income$value > quartiles[3] & middle_income$value <= quartiles [4], ]
middle_income_4 = middle_income[middle_income$value > quartiles[4] & middle_income$value <= quartiles [5],]

#merge subsets into 1 dataframe
merged_subgroup = rbind(middle_income_1, middle_income_2, middle_income_3, middle_income_4)
 #add a column to indicate the quartile
merged_subgroup$quartile = factor(rep(c("Bottom", "Second","Third", "Top"), times = c(nrow(middle_income_1), nrow (middle_income_2), nrow(middle_income_3), nrow(middle_income_4))))

boxplot_data = merge(subgroup_total,
                     merged_subgroup, by = "state")

#print the quartiles
quartiles = quantile(merged_subgroup$value, probs = c(0.25, 0.5, 0.75))
print(paste("First quartile (Q1):", quartiles[1]))
print(paste("Second quartile (Q2):", quartiles[2]))
print(paste("Third quartile (Q3):", quartiles[3]))

#plot the 4 boxplots in 1 plot

library(ggplot2)

ggplot(boxplot_data, aes(x = factor(quartile), y = value.x)) + 
  geom_boxplot()+
  labs(x = "Quartiles in Proportion State Population, making $25k to $49,999", y = "Percentages Point Total Uninsured U.S.", title = "Quartile Analaysis of Uninsured Share among Middle-Income in the US, 2017") + 
  theme_classic()
```

**Event Analysis Visualization, The Impact of The Implementation of Work Requirements for Medicaid in Arkansas, 2018**

-   describe here what happened in Arkansas in 2018 and why the wanted to implement the work requirements and what these requirements were. and that Arkansas is the treatment group and the other 43 states are the control group and the ware analysing if the implementation had significant impact on the uninsured rates compared to the change in the control group which is the weighted average of the uninsured rates.

```{r visualization event, include=FALSE, fig.show = "asis"}
#Here I am making a sub dataframe for the event analysis of the implementation of stricter conditions for Medicaid in 2018 in Arkansas

#sub dataframe Arkansas share uninsured for all the years
arkansas_event =  all_states %>%
  filter(state == "Arkansas", series == "Uninsured.Share") %>%
  select(state,year, value)

#deleting rows
arkansas_event = arkansas_event[-c(2:9),]
arkansas_event = arkansas_event[-c(3:10),]
arkansas_event = arkansas_event[-c(4:11),]
arkansas_event = arkansas_event[-c(5:12),]
arkansas_event = arkansas_event[-c(6:13),]
arkansas_event = arkansas_event[-c(7:14),]
arkansas_event = arkansas_event[-c(8:15),]
arkansas_event = arkansas_event[-c(9:16),]
arkansas_event = arkansas_event[-c(10:17),]
arkansas_event = arkansas_event[-c(11:18),]
arkansas_event = arkansas_event[-c(12:19),]
arkansas_event = arkansas_event[-c(13:20),]
arkansas_event = arkansas_event[-c(14:21),]

View(arkansas_event)

#Here I compute the weighted average of the other 43 states excluding Arkansas
totals = all_states_100 %>%
  filter((str_detect(income_group, "Total civilian noninstitutionalized population")|str_detect(income_group, "Total population")) & state!= "Arkansas") %>%
  select(year, state,total = value, series)

uninsured_share = all_states_100 %>%
  filter(series == "Uninsured.Share" & state!="Arkansas") %>%
  select(year, state, uninsured_share = value)

weighted_avg_2 = totals %>%
  inner_join(uninsured_share, by = c("year", "state"))%>%
  group_by(year)%>%
  summarise(weighted_avg = sum(uninsured_share * total, na.rm = TRUE)/ sum(total, na.rm = TRUE))

#plotting the event
library(ggplot2)

#adding a vertical line at the specific year

event_year = 2018
#The subsets we must use are weighted_avg_2 and arkansas_event

ggplot() + geom_line(data = weighted_avg_2, aes(x = year, y = weighted_avg, color = "Weighted Average"), size = 1) + 
  geom_line(data = arkansas_event, 
            aes(x = year, y = value, color = "Arkansas Share Uninsured"),
            size = 1
            ) + 
  labs(x = "Year", y = "Percentages Point of the Uninsured Population", title = "AR Medicaid Work Requirements, 2018") + 
  scale_color_manual(values = c("Weighted Average" = "darkblue", "Arkansas Share Uninsured" = "darkred")) +
  theme_classic() + geom_vline (xintercept = event_year, color = "black") + annotate("text", x = event_year + 0.3 , y = max(c(weighted_avg_2$weighted_avg, arkansas_event$value)),
                                                                                     label = "Implementation of Requirements in 2018", hjust = 0, size =3, fontface = "bold") +
  scale_x_continuous(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2021,2022,2023)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## 3.2 Identify Trends and Patterns

# Part 4 – Communicate Findings

## 4.1 Summarize Key Insights

Our analysis highlights systemic inequalities in access to health insurance. Despite overall improvements since the ACA, millions remain uninsured.

## 4.2 Propose Solutions or Policy Recommendations

Potential solutions include:

-   Expanding Medicaid in all states
-   Decoupling health insurance from employment
-   Increasing subsidies for marketplace plans

# Appendix

## A.1 References

-   U.S. Census Bureau. (2022). Health Insurance Coverage in the United States.
-   Kaiser Family Foundation. (2023). Key Facts about the Uninsured Population.

## A.2 Session Info
