---
title: "Visualizing the Uninsured: A Data Science Perspective on U.S. Health Coverage"
author: |
  Eduardo Sáenz-Messía Laborda, 2861597 
  Catherina Mikhail, 2847118  
  Charlotte Craenen, 2865572  
  Hugo van As, 2848225
  Indy Pleijter, 2810379
  Julian Zeguers, 2772646  
  Wies Polderman, 2824987 
  Tutorial group 5, group 4 – J.F. Fitzgerald
date: "2025-06-24"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(repos = c(CRAN = "https://cran.rstudio.com/"))

packages <- c("rmarkdown", "tinytex", "rnaturalearth", "rnaturalearthdata", "maps", "tidyverse", "tidyr", "dplyr", "renv", "rnaturalearthhires")

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) {
  install.packages(packages[!installed], repos = "https://cran.rstudio.com/")
}

invisible(lapply(packages, library, character.only = TRUE))

if (!tinytex::is_tinytex()) { 
  tinytex::install_tinytex()
}

if (!require("remotes")) {
  install.packages("remotes")
}
if(!
   require("rnaturalearthhires")){
  remotes:: install_github("reponsci/rnaturalearthhires")
}
library(rnaturalearthhires)
```

**Part 1 – Identify a Social Problem**

## 1.1 Describe the Social Problem

The large number of uninsured people in the United States represents a significant social issue with serious consequences for individuals and the healthcare system. As of 2022, over 27 million Americans—approximately one in twelve people—lacked health insurance coverage (KFF, 2023). This lack of coverage creates substantial barriers to healthcare access and has wide-ranging effects on both personal and public health.

For individuals without insurance, the consequences are often severe. Uninsured people are much more likely to delay or avoid necessary medical care due to cost concerns, which leads to worse health outcomes, preventable emergency room visits, and higher rates of preventable deaths. The financial impact is equally serious, as uninsured individuals must pay full price for medical services, often resulting in medical debt, damaged credit, and long-term financial hardship (Davis, 2007).

The problem extends beyond individual impacts to affect the entire healthcare system. When uninsured patients cannot pay for their care, hospitals and clinics must absorb these costs as uncompensated care. This financial burden strains healthcare providers and ultimately increases costs for everyone else through higher insurance premiums and taxes. As Davis (2007) explains, the uninsured population includes "our neighbors, co-workers, and family members," making this a community-wide issue that affects society as a whole.

## 1.2 Provide Background on the Problem

The Affordable Care Act (ACA), enacted in 2010, significantly expanded health insurance coverage through Medicaid expansion and subsidized insurance marketplaces. However, millions of Americans remain uninsured despite these reforms. According to the Kaiser Family Foundation, most uninsured individuals today are low-income workers, disproportionately people of color, and residents of states that chose not to expand Medicaid under the ACA (KFF, 2023).

A major factor contributing to this problem is the employment-based nature of the U.S. health insurance system. Many uninsured individuals work in jobs that do not offer health benefits, including part-time, temporary, or service sector positions. These workers often earn too much to qualify for traditional Medicaid but too little to afford private insurance, creating what Davis (2007) describes as falling "through the cracks" of the current system.

Geographic disparities also play a significant role. States that expanded Medicaid under the ACA have substantially lower uninsured rates than non-expansion states, creating uneven access to coverage across the country. This policy variation has resulted in particularly high uninsured rates in many Southern and some Western states.

The persistence of uninsurance reflects broader structural challenges in American healthcare, including affordability, accessibility, and policy gaps that leave vulnerable populations without coverage options.

# Part 2 – Describe and Acquire Data

## 2.1 Describe the Data sets

This analysis utilizes two primary datasets to examine patterns in health insurance coverage and their relationship to economic conditions across U.S. states.

**U.S. Census Bureau Data** The main data set comes from the U.S. Census Bureau's American Community Survey (ACS) 1-Year Estimates, specifically the table "Selected Characteristics of the Uninsured in the United States." This data set covers the period from 2010 to 2023, excluding 2020 due to the absence of 1-year estimates for that year.

Eight smaller states and territories were excluded from the analysis—Delaware, District of Columbia, Hawaii, Puerto Rico, North Dakota, Rhode Island, Vermont, and Wyoming—because they are not consistently included in the 1-year estimates. The 1-year estimates were chosen over 5-year estimates because they provide more timely data that can capture short-term changes from economic shifts or policy reforms.

The ACS data set provides state-level information on the civilian non-institutionalized population, including total population counts, number of uninsured individuals, and various demographic and socioeconomic characteristics. For this analysis, we focused on total population figures and household income data (adjusted for inflation) to calculate uninsured rates and understand the economic context.

**Bureau of Economic Analysis Data** Economic data comes from the Bureau of Economic Analysis (BEA) table SAGDP1 – State Annual Gross Domestic Product Summary. This data set provides real GDP figures in millions of chained 2017 dollars for each state from 1997 through 2024. Using inflation-adjusted GDP enables consistent comparisons over time and across different states.

**Data Integration** These two datasets together provide a comprehensive view of how health insurance coverage has changed across the United States over time and how coverage patterns may relate to state-level economic conditions. The combination allows for analysis of both temporal trends and cross-state variations in uninsured rates.

## 2.2 Import and Prepare the Data set

**First data set:** **"Selected Characteristics of The Uninsured in the U.S."**

-   Here we imported the first data set from the U.S. Census Bureau and named them "dataset_uninsured_population\_*year"* from 2010-2023, excl 2020.

```{r import-data, include=FALSE}
# The first dataset is from the dataset "selected characteristics of the uninsured in the US from 2010-2023, exclduing 2020 from the US Census Bureau. These are the ACS one-year estimates.

#2010
ACSST1Y2010_S2702_2025_06_03T123607 <- read_csv("data insured population 2010-2023/ACSST1Y2010.S2702-2025-06-03T123607.csv")
dataset_uninsured_population_2010 = read_csv("data insured population 2010-2023/ACSST1Y2010.S2702-2025-06-03T123607.csv")

#2011
ACSST1Y2011_S2702_2025_06_03T123539 <- read_csv("data insured population 2010-2023/ACSST1Y2011.S2702-2025-06-03T123539.csv")
dataset_uninsured_population_2011 = read_csv("data insured population 2010-2023/ACSST1Y2011.S2702-2025-06-03T123539.csv")

#2012
ACSST1Y2012_S2702_2025_06_03T123441 <- read_csv("data insured population 2010-2023/ACSST1Y2012.S2702-2025-06-03T123441.csv")
dataset_uninsured_population_2012 = read_csv("data insured population 2010-2023/ACSST1Y2012.S2702-2025-06-03T123441.csv")

#2013
ACSST1Y2013_S2702_2025_06_03T123404 <- read_csv("data insured population 2010-2023/ACSST1Y2013.S2702-2025-06-03T123404.csv")
dataset_uninsured_population_2013 = read_csv("data insured population 2010-2023/ACSST1Y2013.S2702-2025-06-03T123404.csv")

#2014
ACSST1Y2014_S2702_2025_06_03T123336 <- read_csv("data insured population 2010-2023/ACSST1Y2014.S2702-2025-06-03T123336.csv")
dataset_uninsured_population_2014 = read_csv("data insured population 2010-2023/ACSST1Y2014.S2702-2025-06-03T123336.csv")

#2015
ACSST1Y2015_S2702_2025_06_03T123312 <- read_csv("data insured population 2010-2023/ACSST1Y2015.S2702-2025-06-03T123312.csv")
dataset_uninsured_population_2015 = read_csv("data insured population 2010-2023/ACSST1Y2015.S2702-2025-06-03T123312.csv")

#2016
ACSST1Y2016_S2702_2025_06_03T123232 <- read_csv("data insured population 2010-2023/ACSST1Y2016.S2702-2025-06-03T123232.csv")
dataset_uninsured_population_2016 = read_csv("data insured population 2010-2023/ACSST1Y2016.S2702-2025-06-03T123232.csv")

#2017
ACSST1Y2017_S2702_2025_06_03T123200 <- read_csv("data insured population 2010-2023/ACSST1Y2017.S2702-2025-06-03T123200.csv")
dataset_uninsured_population_2017 = read_csv("data insured population 2010-2023/ACSST1Y2017.S2702-2025-06-03T123200.csv")

#2018
ACSST1Y2018_S2702_2025_06_03T123128 <- read_csv("data insured population 2010-2023/ACSST1Y2018.S2702-2025-06-03T123128.csv")
dataset_uninsured_population_2018 = read_csv("data insured population 2010-2023/ACSST1Y2018.S2702-2025-06-03T123128.csv")

#2019
ACSST1Y2019_S2702_2025_06_03T123018 <- read_csv("data insured population 2010-2023/ACSST1Y2019.S2702-2025-06-03T123018.csv")
dataset_uninsured_population_2019 = read_csv("data insured population 2010-2023/ACSST1Y2019.S2702-2025-06-03T123018.csv")

#2021
ACSST1Y2021_S2702_2025_06_03T122952 <- read_csv("data insured population 2010-2023/ACSST1Y2021.S2702-2025-06-03T122952.csv")
dataset_uninsured_population_2021 = read_csv("data insured population 2010-2023/ACSST1Y2021.S2702-2025-06-03T122952.csv")

#2022
ACSST1Y2022_S2702_2025_06_03T122917 <- read_csv("data insured population 2010-2023/ACSST1Y2022.S2702-2025-06-03T122917.csv")
dataset_uninsured_population_2022 = read_csv("data insured population 2010-2023/ACSST1Y2022.S2702-2025-06-03T122917.csv")

#2023
ACSST1Y2023_S2702_2025_06_04T165512 <- read_csv("data insured population 2010-2023/ACSST1Y2023.S2702-2025-06-04T165512.csv")
dataset_uninsured_population_2023 = read_csv("data insured population 2010-2023/ACSST1Y2023.S2702-2025-06-04T165512.csv")

```

**Second Data set: "SAGDP1 State Annual Gross Domestic Product Summary", Statistic: "Real GDP (millions of chained 2017 dollars)"**

-   Here we imported the second data set from the Bureau of Economic Analysis of the annual GDP of the states from 2009-2023 and named it "table_GDP"

```{r import second dataset, figure.show="asis", include=F}
#The second dataset is from the Bureau of Economic Analysis and is the annual real GDP of the 44 states we include in our analysis from the year 2009-2023, excluding 2020
annual_gdp_Table_4_ <- read_csv("data GDP/annual gdp - Table (4).csv")

#renaming the table
table_GDP = annual_gdp_Table_4_
```

## Data cleaning: Data sets of Uninsured Population and The Creation of The New Variable: The Uninsured Share

To prepare the data for analysis, we began by removing all margins of error from the original Census data sets. While margins of error are important for assessing statistical precision, they were not necessary for our purposes, which focused on comparing overall trends in health insurance coverage across states and years. Removing them helped simplify the data set and reduce complexity without compromising our analytic goals.

A more structural challenge involved the inclusion of smaller states and territories. Since 2016, the Census Bureau’s 1-year estimates no longer cover certain low-population areas, meaning they are missing from some years in our timespan. To ensure consistency across time, we decided to exclude eight such places entirely from our analysis. These were Delaware, the District of Columbia, Hawaii, Puerto Rico, North Dakota, Rhode Island, Vermont, and Wyoming. By removing them, we ensured that our data set included only states that were present in every applicable year from 2010 to 2023 (with the exception of 2020, when no data was published due to the COVID-19 pandemic).

After resolving issues of consistency and coverage, we turned our attention to the structure of the data itself. Each annual data set included a wide range of metadata—demographic and socioeconomic characteristics that, while valuable for other types of research, were not essential to our primary analysis of insurance coverage trends. We removed variables related to age, sex, race and Hispanic or Latino origin, nativity and U.S. citizenship status, disability status, residence one year ago, educational attainment, employment status, work experience, civilian non-institutionalized workers aged 16 and over, earnings in the past 12 months, and the ratio of income to the poverty level in the past 12 months. From all this, we retained only the total civilian non-institutionalized population and household income (adjusted for inflation), which were directly relevant to our questions about insurance and economic context.

To ensure the data was ready for calculations, we then converted all numeric values that had been stored as character strings—often due to formatting like commas—into actual numeric variables. This was a necessary step for performing any reliable mathematical operations.

One of the most important transformations in our cleaning process was the creation of a new variable: uninsured share. The original data provided the total population and the total number of uninsured individuals for each state and year. To make this more interpretable and comparable, we calculated the uninsured share by dividing the number of uninsured people by the total population and multiplying the result by 100. This gave us a percentage that reflects the proportion of each state’s population without health insurance. Expressing the data this way allows for easier comparisons across time and across states, regardless of differences in population size. It helps illuminate broader patterns in insurance coverage that would be difficult to spot using absolute numbers alone.

After all these cleaning and transformation steps, we combined the individual yearly data sets into a single, comprehensive data set covering the period from 2010 to 2023 (excluding 2020). This final merged file allowed us to analyze long-term trends and relationships in a consistent and structured way.

```{r datacleaning of the yearly datasets Uninsured Population 2010,2023, include=FALSE}
# This is me cleaning out the data insured population from 2010-2016 getting out the margin of error
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2010)))]

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2011)))]

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2012)))]

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2013)))]

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2014)))]

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2015)))]

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2016)))]

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2017)))]

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2018)))]

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2019)))]

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2021)))]

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2022)))]

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2023)))]

###############################################################################

#This is me cleaning out the columns of Delaware
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2010)))]


dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2011)))]


dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2012)))]


dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2013)))]


dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2014)))]


dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2015)))]


dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2016)))]


dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2017)))]


dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2018)))]


dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2019)))]


dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2021)))]


dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2022)))]


dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2023)))]


#This is me cleaning out the columns of District of Columbia
dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2011)))]


dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2012)))]


dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2013)))]


dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2014)))]


dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2015)))]


dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)

#This is me cleaning out the columns of Hawaii
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of North Dakota
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Puerto Rico
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Rhode Island
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Vermont
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Wyoming
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)

###############################################################################

#Now we are cleaning out the unnecessary metadata in the rows
#this is me cleaning out the rows with unnecessary metadata and keeping household income levels and the total population numbers
dataset_uninsured_population_2010 <- dataset_uninsured_population_2010[-c(2:104),]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 <- dataset_uninsured_population_2011[-c(2:104),]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 <- dataset_uninsured_population_2012[-c(2:104),]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 <- dataset_uninsured_population_2013[-c(2:104),]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 <- dataset_uninsured_population_2014[-c(2:104),]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 <- dataset_uninsured_population_2015[-c(2:104),]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 <- dataset_uninsured_population_2016[-c(2:104),]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 <- dataset_uninsured_population_2017[-c(2:104),]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 <- dataset_uninsured_population_2018[-c(2:104),]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 <- dataset_uninsured_population_2019[-c(2:104),]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 <- dataset_uninsured_population_2021[-c(2:104),]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 <- dataset_uninsured_population_2022[-c(2:104),]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 <- dataset_uninsured_population_2023[-c(2:104),]
View(dataset_uninsured_population_2023)

dataset_uninsured_population_2010 <- dataset_uninsured_population_2010[-c(10:17),]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 <- dataset_uninsured_population_2011[-c(10:17),]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 <- dataset_uninsured_population_2012[-c(10:17),]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 <- dataset_uninsured_population_2013[-c(10:17),]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 <- dataset_uninsured_population_2014[-c(10:17),]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 <- dataset_uninsured_population_2015[-c(10:17),]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 <- dataset_uninsured_population_2016[-c(10:17),]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 <- dataset_uninsured_population_2017[-c(10:17),]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 <- dataset_uninsured_population_2018[-c(10:17),]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 <- dataset_uninsured_population_2019[-c(10:17),]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 <- dataset_uninsured_population_2021[-c(10:17),]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 <- dataset_uninsured_population_2022[-c(10:17),]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 <- dataset_uninsured_population_2023[-c(10:17),]
View(dataset_uninsured_population_2023)

################################################################################

#As the values in these yearly datasets are stored as characters, we are now going to coerce them into numeric values so that we can create a new variable afterwards.
#The values in the datasets of the uninsured population are stored as characters. Here I am going to get rid of "" so I can store them as numeric.

# Keep the first column as-is
first_col <- dataset_uninsured_population_2010[[1]]

# Clean and convert the rest of the columns
# Keep the first column as-is
first_col <- dataset_uninsured_population_2010[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2010_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2010[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2010_cleaned <- cbind(first_col, dataset_uninsured_population_2010_numeric)

#rename the first column
colnames(dataset_uninsured_population_2010_cleaned)[1] <- colnames(dataset_uninsured_population_2010)[1]

View(dataset_uninsured_population_2010_cleaned)
class(dataset_uninsured_population_2010_cleaned)
class(dataset_uninsured_population_2010_cleaned[[2]])




# Keep the first column as-is
first_col <- dataset_uninsured_population_2011[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2011_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2011[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2011_cleaned <- cbind(first_col, dataset_uninsured_population_2011_numeric)

#rename the first column
colnames(dataset_uninsured_population_2011_cleaned)[1] <- colnames(dataset_uninsured_population_2011)[1]

View(dataset_uninsured_population_2011_cleaned)
class(dataset_uninsured_population_2011_cleaned)
class(dataset_uninsured_population_2011_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2012[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2012_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2012[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2012_cleaned <- cbind(first_col, dataset_uninsured_population_2012_numeric)

#rename the first column
colnames(dataset_uninsured_population_2012_cleaned)[1] <- colnames(dataset_uninsured_population_2012)[1]

View(dataset_uninsured_population_2012_cleaned)
class(dataset_uninsured_population_2012_cleaned)
class(dataset_uninsured_population_2012_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2013[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2013_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2013[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2013_cleaned <- cbind(first_col, dataset_uninsured_population_2013_numeric)

#rename the first column
colnames(dataset_uninsured_population_2013_cleaned)[1] <- colnames(dataset_uninsured_population_2013)[1]

View(dataset_uninsured_population_2013_cleaned)
class(dataset_uninsured_population_2013_cleaned)
class(dataset_uninsured_population_2013_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2014[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2014_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2014[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2014_cleaned <- cbind(first_col, dataset_uninsured_population_2014_numeric)

#rename the first column
colnames(dataset_uninsured_population_2014_cleaned)[1] <- colnames(dataset_uninsured_population_2014)[1]

View(dataset_uninsured_population_2014_cleaned)
class(dataset_uninsured_population_2014_cleaned)
class(dataset_uninsured_population_2014_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2015[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2015_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2015[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2015_cleaned <- cbind(first_col, dataset_uninsured_population_2015_numeric)

#rename the first column
colnames(dataset_uninsured_population_2015_cleaned)[1] <- colnames(dataset_uninsured_population_2015)[1]

View(dataset_uninsured_population_2015_cleaned)
class(dataset_uninsured_population_2015_cleaned)
class(dataset_uninsured_population_2015_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2016[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2016_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2016[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2016_cleaned <- cbind(first_col, dataset_uninsured_population_2016_numeric)

#rename the first column
colnames(dataset_uninsured_population_2016_cleaned)[1] <- colnames(dataset_uninsured_population_2016)[1]

View(dataset_uninsured_population_2016_cleaned)
class(dataset_uninsured_population_2016_cleaned)
class(dataset_uninsured_population_2016_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2017[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2017_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2017[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2017_cleaned <- cbind(first_col, dataset_uninsured_population_2017_numeric)

#rename the first column
colnames(dataset_uninsured_population_2017_cleaned)[1] <- colnames(dataset_uninsured_population_2017)[1]

View(dataset_uninsured_population_2017_cleaned)
class(dataset_uninsured_population_2017_cleaned)
class(dataset_uninsured_population_2017_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2018[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2018_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2018[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2018_cleaned <- cbind(first_col, dataset_uninsured_population_2018_numeric)

#rename the first column
colnames(dataset_uninsured_population_2018_cleaned)[1] <- colnames(dataset_uninsured_population_2018)[1]

View(dataset_uninsured_population_2018_cleaned)
class(dataset_uninsured_population_2018_cleaned)
class(dataset_uninsured_population_2018_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2019[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2019_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2019[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2019_cleaned <- cbind(first_col, dataset_uninsured_population_2019_numeric)

#rename the first column
colnames(dataset_uninsured_population_2019_cleaned)[1] <- colnames(dataset_uninsured_population_2019)[1]

View(dataset_uninsured_population_2019_cleaned)
class(dataset_uninsured_population_2019_cleaned)
class(dataset_uninsured_population_2019_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2021[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2021_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2021[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2021_cleaned <- cbind(first_col, dataset_uninsured_population_2021_numeric)

#rename the first column
colnames(dataset_uninsured_population_2021_cleaned)[1] <- colnames(dataset_uninsured_population_2021)[1]

View(dataset_uninsured_population_2021_cleaned)
class(dataset_uninsured_population_2021_cleaned)
class(dataset_uninsured_population_2021_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2022[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2022_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2022[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2022_cleaned <- cbind(first_col, dataset_uninsured_population_2022_numeric)

#rename the first column
colnames(dataset_uninsured_population_2022_cleaned)[1] <- colnames(dataset_uninsured_population_2022)[1]

View(dataset_uninsured_population_2022_cleaned)
class(dataset_uninsured_population_2022_cleaned)
class(dataset_uninsured_population_2022_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2023[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2023_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2023[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2023_cleaned <- cbind(first_col, dataset_uninsured_population_2023_numeric)

#rename the first column
colnames(dataset_uninsured_population_2023_cleaned)[1] <- colnames(dataset_uninsured_population_2023)[1]

View(dataset_uninsured_population_2023_cleaned)
class(dataset_uninsured_population_2023_cleaned)
class(dataset_uninsured_population_2023_cleaned[[2]])

################################################################################

#Now that the values in the yearly datasets are stored as numeric values but the first column as the first column contains the names of the states, we can now create are first new variable, which is the uninsured share:

#Here I am trying to compute and create a new variable, which is the share of uninsured population per state and per year

# Copy original data 2010
df_pop_2010 <- dataset_uninsured_population_2010_cleaned
col_names <- colnames(df_pop_2010)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2010[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2010[[total_col]][1]) && df_pop_2010[[total_col]][1] != 0) {
      df_pop_2010[[share_col]][1] <- df_pop_2010[[uninsured_col]][1] / df_pop_2010[[total_col]][1]
    } else {
      df_pop_2010[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2010 <- df_pop_2010[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2010), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2010[1, share_cols])

View(df_pop_2010)

#################################################################################

# Copy original data 2011
df_pop_2011 <- dataset_uninsured_population_2011_cleaned
col_names <- colnames(df_pop_2011)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2011[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2011[[total_col]][1]) && df_pop_2011[[total_col]][1] != 0) {
      df_pop_2011[[share_col]][1] <- df_pop_2011[[uninsured_col]][1] / df_pop_2011[[total_col]][1]
    } else {
      df_pop_2011[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2011 <- df_pop_2011[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2011), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2011[1, share_cols])

View(df_pop_2011)


###############################################################################


# Copy original data 2012
df_pop_2012 <- dataset_uninsured_population_2012_cleaned
col_names <- colnames(df_pop_2012)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2012[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2012[[total_col]][1]) && df_pop_2012[[total_col]][1] != 0) {
      df_pop_2012[[share_col]][1] <- df_pop_2012[[uninsured_col]][1] / df_pop_2012[[total_col]][1]
    } else {
      df_pop_2012[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2012 <- df_pop_2012[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2012), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2012[1, share_cols])


###############################################################################


# Copy original data 2013
df_pop_2013 <- dataset_uninsured_population_2013_cleaned
col_names <- colnames(df_pop_2013)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2013[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2013[[total_col]][1]) && df_pop_2013[[total_col]][1] != 0) {
      df_pop_2013[[share_col]][1] <- df_pop_2013[[uninsured_col]][1] / df_pop_2013[[total_col]][1]
    } else {
      df_pop_2013[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2013 <- df_pop_2013[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2013), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2013[1, share_cols])


################################################################################


# Copy original data 2014
df_pop_2014 <- dataset_uninsured_population_2014_cleaned
col_names <- colnames(df_pop_2014)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2014[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2014[[total_col]][1]) && df_pop_2014[[total_col]][1] != 0) {
      df_pop_2014[[share_col]][1] <- df_pop_2014[[uninsured_col]][1] / df_pop_2014[[total_col]][1]
    } else {
      df_pop_2014[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2014 <- df_pop_2014[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2014), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2014[1, share_cols])


################################################################################


# Copy original data 2015
df_pop_2015 <- dataset_uninsured_population_2015_cleaned
col_names <- colnames(df_pop_2015)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2015[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2015[[total_col]][1]) && df_pop_2015[[total_col]][1] != 0) {
      df_pop_2015[[share_col]][1] <- df_pop_2015[[uninsured_col]][1] / df_pop_2015[[total_col]][1]
    } else {
      df_pop_2015[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2015 <- df_pop_2015[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2015), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2015[1, share_cols])


###############################################################################


# Copy original data 2016
df_pop_2016 <- dataset_uninsured_population_2016_cleaned
col_names <- colnames(df_pop_2016)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2016[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2016[[total_col]][1]) && df_pop_2016[[total_col]][1] != 0) {
      df_pop_2016[[share_col]][1] <- df_pop_2016[[uninsured_col]][1] / df_pop_2016[[total_col]][1]
    } else {
      df_pop_2016[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2016 <- df_pop_2016[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2016), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2016[1, share_cols])

################################################################################

#something went wrong here with 2017 and 2018, so the function that worked is much more below as the column names in the years of 2017-2018 were named differently than the years before

# Copy original data 2017
df_pop_2017 <- dataset_uninsured_population_2017_cleaned
col_names <- colnames(df_pop_2017)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2017[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2017[[total_col]][1]) && df_pop_2017[[total_col]][1] != 0) {
      df_pop_2017[[share_col]][1] <- df_pop_2017[[uninsured_col]][1] / df_pop_2017[[total_col]][1]
    } else {
      df_pop_2017[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2017 <- df_pop_2017[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2017), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2017[1, share_cols])

# Copy original data 2018
df_pop_2018<- dataset_uninsured_population_2018_cleaned
col_names <- colnames(df_pop_2018)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2018[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2018[[total_col]][1]) && df_pop_2018[[total_col]][1] != 0) {
      df_pop_2018[[share_col]][1] <- df_pop_2018[[uninsured_col]][1] / df_pop_2018[[total_col]][1]
    } else {
      df_pop_2018[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2018 <- df_pop_2018[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2018), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2018[1, share_cols])

View(df_pop_2017)
View(df_pop_2016)
View(df_pop_2018)

#I have seen that it did not work for the years of 2017 and 2018 so I have to see what the names of the columns are
grep("Alabama", colnames(dataset_uninsured_population_2017_cleaned), value = TRUE)
grep("Alabama", colnames(dataset_uninsured_population_2018_cleaned), value = TRUE)

#Here I am starting again from 2017

# Make a working copy of your data 2017
df_pop_2017 <- dataset_uninsured_population_2017_cleaned
col_names <- colnames(df_pop_2017)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2017[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2017[[total_col]][1]
    uninsured_val <- df_pop_2017[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2017[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2017 <- df_pop_2017[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2017), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2017[1, share_cols])

#now it worked and I can work further on 2018-2023 except for 2020 as we do not have the data of that

# Make a working copy of your data 2018
df_pop_2018 <- dataset_uninsured_population_2018_cleaned
col_names <- colnames(df_pop_2018)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2018[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2018[[total_col]][1]
    uninsured_val <- df_pop_2018[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2018[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2018 <- df_pop_2018[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2018), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2018[1, share_cols])


################################################################################


# Make a working copy of your data 2019
df_pop_2019 <- dataset_uninsured_population_2019_cleaned
col_names <- colnames(df_pop_2019)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2019[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2019[[total_col]][1]
    uninsured_val <- df_pop_2019[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2019[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2019 <- df_pop_2019[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2019), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2019[1, share_cols])


################################################################################


# Make a working copy of your data 2021
df_pop_2021 <- dataset_uninsured_population_2021_cleaned
col_names <- colnames(df_pop_2021)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2021[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2021[[total_col]][1]
    uninsured_val <- df_pop_2021[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2021[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2021 <- df_pop_2021[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2021), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2021[1, share_cols])

###############################################################################


# Make a working copy of your data 2022
df_pop_2022 <- dataset_uninsured_population_2022_cleaned
col_names <- colnames(df_pop_2022)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2022[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2022[[total_col]][1]
    uninsured_val <- df_pop_2022[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2022[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2022 <- df_pop_2022[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2022), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2022[1, share_cols])


################################################################################

# Make a working copy of your data 2023
df_pop_2023 <- dataset_uninsured_population_2023_cleaned
col_names <- colnames(df_pop_2023)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2023[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2023[[total_col]][1]
    uninsured_val <- df_pop_2023[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2023[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2023 <- df_pop_2023[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2023), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2023[1, share_cols])

################################################################################

#Now we have the new variable in the seperate datasets and now we are going to merge these yearly datasets into 1 dataset

# Define years
years_early <- 2010:2016
years_late <- c(2017:2019, 2021:2023)

# Function to process one year of data
process_df <- function(df, year, is_early) {
  label_col <- names(df)[1]
  state_cols <- names(df)[!names(df) %in% c(label_col, "Unnamed: 0")]
  states <- unique(str_extract(state_cols, "^[^.]+"))
  
  # Set correct suffixes depending on year
  if (is_early) {
    pop_suffix <- "..Total.Civilian.Noninstitutionalized.Population..Estimate"
    unins_suffix <- "..Uninsured.Population..Estimate"
    share_suffix <- "..Uninsured.Share"
  } else {
    pop_suffix <- "..Total..Estimate"
    unins_suffix <- "..Total.Uninsured..Estimate"
    share_suffix <- "..Uninsured.Share"
  }
  
  # Extract for each state
  rows <- lapply(states, function(state) {
    pop_col <- paste0(state, pop_suffix)
    unins_col <- paste0(state, unins_suffix)
    share_col <- paste0(state, share_suffix)
    
    # Get only total population row
    total_row <- df %>%
      filter(str_trim(.data[[label_col]]) == "Total population")
    
    if (nrow(total_row) == 0 || !(pop_col %in% names(df))) return(NULL)
    
    tibble(
      state = state,
      year = year,
      total_population = as.numeric(total_row[[pop_col]]),
      total_uninsured = as.numeric(total_row[[unins_col]]),
      uninsured_share = as.numeric(total_row[[share_col]])
    )
  })
  
  bind_rows(rows)
}

# Process all early years
dfs_early <- mget(paste0("df_pop_", years_early))
panel_early <- map2_dfr(dfs_early, years_early, ~process_df(.x, .y, is_early = TRUE))

# Process all late years
dfs_late <- mget(paste0("df_pop_", years_late))
#panel_late <- map2_dfr(dfs_late, years_late, ~process_df(.x, .y, is_early = FALSE))

early_states <- bind_rows(
  dfs_early, .id = "year") %>%
  pivot_longer(
    -c(year, "Label (Grouping)"),
    names_to = "tmp",
    values_to = "value") %>%
  separate(tmp,
           into = c("state", "series", "measure"),
           sep = "\\.\\.",
           remove = TRUE)

late_states <- bind_rows(
  dfs_late, .id = "year") %>%
  pivot_longer(
    -c(year, "Label (Grouping)"),
    names_to = "tmp",
    values_to = "value") %>%
  separate(tmp,
           into = c("state", "series", "measure"),
           sep = "\\.\\.",
           remove = TRUE)
# View

all_states <- rbind(late_states, early_states)

head(all_states)
###############################################################################

#cleaning out df_pop from the column of years to only remain the year numbers

unique(all_states$year)

all_states$year <- str_remove(all_states$year, "df_pop_")

#Now I am coercing the string into numeric so plotting becomes easier

all_states$year <- as.numeric(all_states$year)
str(all_states)

################################################################################
#Here I am multiplying the uninsured share by 100, such that they are representing percentages
all_states_100 = all_states %>%
  mutate(value = 
           ifelse(grepl("Uninsured.Share",
                        series), value*100, value))

#This is the dataset for the Uninsured Population we can use for the visualizations
#renaming a column to make it easier "Label (Grouping)" to "income_group"

library(dplyr)
all_states_100 = all_states_100 %>%
  rename(income_group = `Label (Grouping)`)
```

## Data cleaning: Annual GDP Table and The Creation of The New Variable: Annual Real GDP Growth per year and state

For the economic data from the Bureau of Economic Analysis, we worked with the *SAGDP1 – State Annual Gross Domestic Product Summary* table, which reports real GDP per state in millions of chained 2017 dollars. Before analysis, we cleaned this data set to align it with the structure and timeframe of our health insurance data. First, we removed all columns corresponding to years outside our scope—specifically, 1997 through 2008, as well as 2020 and 2024. We also filtered out states and regions that were not included in our main data set or that represented aggregate regions rather than individual states. Specifically, we excluded Delaware, District of Columbia, Hawaii, North Dakota, Puerto Rico, Vermont, Wyoming, and Rhode Island, as well as the regional aggregates: New England, Mideast, Great Lakes, Plains, Southeast, Southwest, Rocky Mountain, and Far West. Also here, the values were stored as "double", so we coerced the values to numeric, so that we are able to compute our new variable. After these adjustments, we renamed the cleaned data set `table_GDP` for use in our analysis.

From this cleaned table, we created a new variable: annual real GDP growth, calculated for each state and year by measuring the year-over-year percentage change in real GDP. This variable allowed us to capture how quickly or slowly each state’s economy was growing over time. Including GDP growth in our analysis adds important context: it helps us explore whether changes in economic performance are linked to shifts in health insurance coverage. For example, we can examine whether periods of economic expansion correspond with improvements in coverage rates. By combining this with our uninsurance data, we gain a more nuanced understanding of how economic conditions may influence access to healthcare.

```{r include=FALSE}
#Here I am making a new variable which is the GDP growth rate of the states

#renaming the table
table_GDP = annual_gdp_Table_4_

#cleaning out the table of the years 1997:2008, 2020 and 2024 and removing the states/regions that are not included in our main data set

table_GDP = annual_gdp_Table_4_[, !(names(annual_gdp_Table_4_) %in% c("1997", "1998","1999","2000", "2001","2002","2003","2004","2005","2006","2007","2008","2020","2024"))]

table_GDP = table_GDP[-c(9,10,13,36,41,47,52,54,55,56,57,58,59,60),]
table_GDP = table_GDP[-c(1),]
table_GDP = table_GDP[-c(45),]

#class coercion of double to numeric of all columns but the first column
table_GDP[, -1] = lapply(table_GDP[, -1], as.numeric)

#checking the class
sapply(table_GDP, class)

#yes, all the numbers are now numeric instead of the first column with the names of the states


# Pivot longer
gdp_long <- table_GDP %>%
  pivot_longer(
    cols = -GeoName,
    names_to = "year",
    values_to = "gdp"
  ) %>%
  mutate(year = as.integer(year))

View(table_GDP)
View(gdp_long)

# Calculate GDP growth rate
gdp_growth <- gdp_long %>%
  group_by(GeoName) %>%
  arrange(year) %>%
  mutate(growth_rate = (gdp - lag(gdp)) / lag(gdp) * 100)

View(gdp_growth)

#delete rows from 2009
gdp_growth = gdp_growth[-c(1:45), ]
```

# Part 3 – Visualize and Analyze the Data

## 3.1 Create Initial Visualizations

**Spatial Variation Visualization, U.S. Share of The Uninsured Population in 2023**

-   describe here that we mapped here the spatial variation of the uninsured population of the states of which we have the ACS one year estimates of and that we first created a subset of 2023 of the big dataset all_states_100 and that we plotted Alaska separately otherwise the whole map would be too small to visualize.

```{r visualize spatial, echo = FALSE, fig.cap = ""}
#Here I am making a sub dataframe for the spatial visualization of 2023 of all the states with available data

spatial_df = all_states %>%
  filter(year == 2023, series == "Uninsured.Share") %>%
  select(state, value)

#deleting unnecessary rows
spatial_df = spatial_df[-c(45:396),]

#Here I am trying to plot the spatial subset

#getting the US map
us_map = map_data("state")

#clean up states names in both datasets
us_map$region = tolower(us_map$region)
spatial_df$state = tolower(spatial_df$state)

#join the map data with the subset data
us_map = left_join(us_map, spatial_df, by = c("region" = "state"))

#plot the map
ggplot(us_map, aes(x = long, y = lat, group = group, fill = value)) + 
  geom_polygon (color = "black") +
  scale_fill_gradient(low = "lightblue", high = "red", na.value = "white")+
  ggtitle("Share of Uninsured Population in the US, 2023")+
  labs(fill = "(%) Share of Uninsured Population") + 
  theme_void() 

#I still need the state Alaska but the package maps does not include the state Alaska

# I am trying it here again including Alaska with another package

library(rnaturalearth)
library(ggplot2)
library(dplyr)

#get US shapefile
us_states = ne_states(country = "United States of America", returnclass = "sf")

#convert state names into lowercase letters such that they match
us_states$name = tolower(us_states$name)

alaska = us_states[us_states$name == "alaska",]

#merge data with shapefile
alaska_data = alaska %>%
  left_join(spatial_df, by = c("name" = "state"))

#plot the map with the data
ggplot(alaska_data) + 
  geom_sf(aes(fill = value)) + 
  theme_void() + 
  scale_fill_gradient( low = "lightblue", high = "red", na.value = "white")+
  theme(legend.position = "none")
```

**Temporal Variation Visualization, Texas's Share of Uninsured Population and Texas's GDP growth rate over time**

-   describe that Texas stood out in our spatial variation analysis and that we wanted to dig deeper into Texas's share of uninsured population over time from 2010 until 2023, except for 2020 and that we wanted to see if economic growth played a role in this.

```{r visualization temporal, echo = FALSE, fig.cap = ""}
#Here I am making a sub dataframe for the temporal visualization with the GDP growth rate of Texas and another subset of the share of uninsured population in the US over the years

#subset Texas share of uninsured population
texas_share = all_states_100 %>%
  filter(state == "Texas",
         year %in% 2010:2023,
         series == "Uninsured.Share")

#cleaning out the NA rows of subset Texas
texas_share = texas_share[-c(2:9),]
texas_share = texas_share[-c(3:10),]
texas_share = texas_share[-c(4:11),]
texas_share = texas_share[-c(5:12),]
texas_share = texas_share[-c(6:13),]
texas_share = texas_share[-c(7:14),]
texas_share = texas_share[-c(8:15),]
texas_share = texas_share[-c(9:16),]
texas_share = texas_share[-c(10:17),]
texas_share = texas_share[-c(11:18),]
texas_share = texas_share[-c(12:19),]
texas_share = texas_share[-c(13:20),]
texas_share = texas_share[-c(14:21),]


#subset GDP growth of texas of the years 2010-2023 except for 2020
library(dplyr)
texas_gdp = gdp_growth %>%
  filter(GeoName == "Texas",
         year %in% 2010:2023) %>%
  select(year, growth_rate)

#Here I will plot the temporal visualization of Texas over the years 2010-2023 except for year 2020

#the subsets I will use are
View(texas_share)
View(texas_gdp)

library(ggplot2)
library(dplyr)

#Add a type column to distinguish between GDP growth rate and Uninsured Share
texas_gdp$type = "GDP growth rate"
texas_share$type = "Uninsured Share"

#rename the columns to match
texas_share = texas_share %>%
  rename(gdp = value)

#combine the 2 dataframes
combined_data_texas = bind_rows(texas_gdp %>%
                                  select(year, gdp = growth_rate, type),
                                texas_share %>%
                                  select(year, gdp, type))

################################################################################

#plotting the lines from the 2 dataframes

ggplot() + 
  geom_line(data = texas_gdp, aes(x = year, y = growth_rate, color = "GDP Growth Rate"), size =1) +
  geom_line(data = texas_share, aes(x = year, y = gdp, color = "Uninsured Share"), size = 1) + 
  labs(title = "GDP growth rate and Uninsured Share in Texas 2010 - 2023, except 2020", x = "Year", y = "Percentage Points", color = "Type") +
  scale_y_continuous(breaks = seq(-2.5,25, by = 2.5), limits = c(-2.5,25)) + 
  scale_x_continuous(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2021,2022,2023)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

**Sub-Group Variation Visualization, The Uninsured Rates of the Middle-Income Group in 2017 in the U.S.**

-   describe that already the KFF (2024) found that the uninsured people in the US are more likely to be low income and that we were wondering what the state of the middle-income group is as they earn too much for Medicaid but not enough for private insurance. 2017 is the year after Trump became president in 2016.

```{r visualization subgroup, echo = FALSE}
#create a subset of data 2017 for total unsinred share in the states

library(dplyr)

subgroup_total = all_states_100 %>%
  filter(year == 2017,
         series == "Uninsured.Share",
         income_group == "Total civilian noninstitutionalized population")
# create a subset data of 2017 and middle income
middle_income = all_states_100[ all_states_100$income_group == ("$25,000 to $49,999"),]
middle_income = middle_income[middle_income$series == "Total.Uninsured",]
#subset that will be used is middle_income of 2017
#calculate quartiles
quartiles = quantile(middle_income$value, probs = c(0, 0.25, 0.5, 0.75, 1))

middle_income_1 = middle_income[middle_income$value >= quartiles[1] & middle_income$value <= quartiles[2],]
middle_income_2 = middle_income[middle_income$value > quartiles[2] & middle_income$value <= quartiles [3],]
middle_income_3 = middle_income[middle_income$value > quartiles[3] & middle_income$value <= quartiles [4], ]
middle_income_4 = middle_income[middle_income$value > quartiles[4] & middle_income$value <= quartiles [5],]

#merge subsets into 1 dataframe
merged_subgroup = rbind(middle_income_1, middle_income_2, middle_income_3, middle_income_4)
 #add a column to indicate the quartile
merged_subgroup$quartile = factor(rep(c("Bottom", "Second","Third", "Top"), times = c(nrow(middle_income_1), nrow (middle_income_2), nrow(middle_income_3), nrow(middle_income_4))))

boxplot_data = merge(subgroup_total,
                     merged_subgroup, by = "state")

#print the quartiles
quartiles = quantile(merged_subgroup$value, probs = c(0.25, 0.5, 0.75))
print(paste("First quartile (Q1):", quartiles[1]))
print(paste("Second quartile (Q2):", quartiles[2]))
print(paste("Third quartile (Q3):", quartiles[3]))

#plot the 4 boxplots in 1 plot

library(ggplot2)

ggplot(boxplot_data, aes(x = factor(quartile), y = value.x)) + 
  geom_boxplot()+
  labs(x = "Quartiles in Proportion State Population, making $25k to $49,999", y = "Percentages Point Total Uninsured U.S.", title = "Uninsured Share among Middle-Income in the US, 2017") + 
  theme_classic()
```

**Event Analysis Visualization, The Impact of The Implementation of Work Requirements for Medicaid in Arkansas, 2018**

-   describe here what happened in Arkansas in 2018 and why the wanted to implement the work requirements and what these requirements were. and that Arkansas is the treatment group and the other 43 states are the control group and the ware analysing if the implementation had significant impact on the uninsured rates compared to the change in the control group which is the weighted average of the uninsured rates.

```{r visualization event, echo = FALSE, fig.cap = ""}
#Here I am making a sub dataframe for the event analysis of the implementation of stricter conditions for Medicaid in 2018 in Arkansas

#sub dataframe Arkansas share uninsured for all the years
arkansas_event =  all_states %>%
  filter(state == "Arkansas", series == "Uninsured.Share") %>%
  select(state,year, value)

#deleting rows
arkansas_event = arkansas_event[-c(2:9),]
arkansas_event = arkansas_event[-c(3:10),]
arkansas_event = arkansas_event[-c(4:11),]
arkansas_event = arkansas_event[-c(5:12),]
arkansas_event = arkansas_event[-c(6:13),]
arkansas_event = arkansas_event[-c(7:14),]
arkansas_event = arkansas_event[-c(8:15),]
arkansas_event = arkansas_event[-c(9:16),]
arkansas_event = arkansas_event[-c(10:17),]
arkansas_event = arkansas_event[-c(11:18),]
arkansas_event = arkansas_event[-c(12:19),]
arkansas_event = arkansas_event[-c(13:20),]
arkansas_event = arkansas_event[-c(14:21),]

View(arkansas_event)

#Here I compute the weighted average of the other 43 states excluding Arkansas
totals = all_states_100 %>%
  filter((str_detect(income_group, "Total civilian noninstitutionalized population")|str_detect(income_group, "Total population")) & state!= "Arkansas") %>%
  select(year, state,total = value, series)

uninsured_share = all_states_100 %>%
  filter(series == "Uninsured.Share" & state!="Arkansas") %>%
  select(year, state, uninsured_share = value)

weighted_avg_2 = totals %>%
  inner_join(uninsured_share, by = c("year", "state"))%>%
  group_by(year)%>%
  summarise(weighted_avg = sum(uninsured_share * total, na.rm = TRUE)/ sum(total, na.rm = TRUE))

#plotting the event
library(ggplot2)

#adding a vertical line at the specific year

event_year = 2018
#The subsets we must use are weighted_avg_2 and arkansas_event

ggplot() + geom_line(data = weighted_avg_2, aes(x = year, y = weighted_avg, color = "Weighted Average"), size = 1) + 
  geom_line(data = arkansas_event, 
            aes(x = year, y = value, color = "Arkansas Share Uninsured"),
            size = 1
            ) + 
  labs(x = "Year", y = "Percentages Point of the Uninsured Population", title = "AR Medicaid Work Requirements, 2018") + 
  scale_color_manual(values = c("Weighted Average" = "darkblue", "Arkansas Share Uninsured" = "darkred")) +
  theme_classic() + geom_vline (xintercept = event_year, color = "black") + annotate("text", x = event_year + 0.3 , y = max(c(weighted_avg_2$weighted_avg, arkansas_event$value)),
                                                                                     label = "Implementation \nRequirements", hjust = 0, size =3, fontface = "bold") +
  scale_x_continuous(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2021,2022,2023)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## 3.2 Identify Trends and Patterns

**Spatial Variation Visualization, U.S. Share of The Uninsured Population in 2023**

The spatial distribution of uninsured rates across the U.S. in 2023 reveals a stark regional divide. States in the South, especially Texas and Florida, report the highest uninsured shares. In contrast, Northeastern and Upper Midwestern states show much lower rates, likely due to more expansive Medicaid programs, higher employer-based coverage, and more generous state-level health initiatives (Bailey et al., 2017).

More nuanced is the case of Western states like Nevada and Arizona, which also exhibit elevated uninsured rates despite having relatively progressive policy environments. This may reflect labor market characteristics—such as a higher prevalence of gig work, part-time jobs, and self-employment—which are often associated with lower access to employer-sponsored insurance (Collins et al., 2020). Furthermore, undocumented immigrant populations, more common in some Western and Southern states, are often excluded from public coverage (KFF, 2023), further contributing to uninsured rates.

The findings support existing literature showing that health coverage in the U.S. is not solely driven by state policy, but is deeply intertwined with employment structures, income levels, and demographic profiles (Davis, 2007; Sommers et al., 2012). Thus, even among states with similar political leanings, uninsured rates can diverge significantly based on labor and population dynamics.

**Temporal Variation Visualization, Texas's Share of Uninsured Population and Texas's GDP growth rate over time**

The line graph illustrates the relationship between GDP growth rate and the uninsured share in Texas from 2010 to 2023 (excluding 2020).

We observe a clear downward trend in the uninsured share, falling from about 23% to 17% over this period. However, despite economic fluctuations—particularly a dip in GDP growth around 2016 and a spike in 2023—the uninsured rate remained relatively high and stable, especially after 2016.

This pattern suggests that economic growth alone doesn't guarantee increased health insurance coverage. Even in years with strong GDP growth, Texas continued to have one of the highest uninsured rates in the country. A likely explanation is Texas’s decision not to expand Medicaid under the ACA, which left many low-income adults without affordable coverage options (KFF, 2023; Davis, 2007). In other words, policy choices, not just economic performance, shape access to insurance.

**Sub-Group Variation Visualization, The Uninsured Rates of the Middle-Income Group in 2017 in the U.S.**

The boxplot shows the total uninsured share per U.S. state in 2023, grouped by quartiles of the proportion of middle-income earners (those earning \$25,000–\$49,999). Importantly, the y-axis reflects the overall uninsured rate, not just for middle-income individuals.

We observe that the second and third quartiles have nearly the same median uninsured rate, suggesting that having more middle-income residents doesn’t clearly correlate with a higher or lower uninsured share. This supports the idea that income alone doesn’t determine insurance coverage—other factors such as cost of living, Medicaid eligibility rules, and state-level policy differences play major roles.

For example, two states may have similar income distributions, but very different uninsured rates due to differences in insurance affordability, healthcare access, or Medicaid expansion status (Davis, 2007; KFF, 2023). This makes it clear that middle-income presence is not a consistent predictor of how many people remain uninsured at the state level.

**Event Analysis Visualization, The Impact of The Implementation of Work Requirements for Medicaid in Arkansas, 2018**

The graph shows the share of uninsured people in Arkansas compared to the national weighted average from 2010 to 2023, excluding Arkansas. Arkansas saw a sharp decline in its uninsured rate after expanding Medicaid under the Affordable Care Act, reaching a low around 2016–2017. However, after implementing work requirements for Medicaid in 2018, the trend reversed: the uninsured rate in Arkansas increased slightly, while the national average stayed low and stable.

This suggests the policy had a negative impact on coverage. Although the goal was to encourage employment, research shows many people lost coverage not because they were unwilling to work, but because of administrative hurdles and confusion about reporting (Sommers et al., 2019). As a result, thousands of eligible people were dropped from Medicaid, increasing the uninsured share.

The work requirements likely influenced the y-axis variable, which is the percentage uninsured, by creating new barriers to maintaining insurance, especially for low-income adults.

# Part 4 – Communicate Findings

## 4.1 Summarize Key Insights

To summarize our key insights, the spatial variation analysis showed us that regional policy choices and labour market structures significantly influence uninsured rates and not just political ideology that influences health care policies. Besides that, we can conclude based on the temporal analysis that economic growth alone does not automatically translate into improved health coverage, especially in states that did not expand Medicaid. Also, our subgroup analysis gave us the insight that income alone does not fully explain the variation in uninsured rates. Factors like the overall cost of living in a state, state policies and insurance access also play major roles. Finally, the event analysis showed us that the implementation of work requirements for Medicaid may reverse coverage gains and disproportionately impact low-income adults who face employment instability.

### 

## 4.2 Propose Solutions or Policy Recommendations

## Key Policy Solutions

Based on the analysis findings, four targeted policy interventions are recommended:

### 1. Medicaid Expansion Incentives

Provide enhanced federal funding to encourage remaining non-expansion states like Texas and Florida to adopt Medicaid expansion. The temporal analysis shows economic growth alone cannot close coverage gaps without policy action.

### 2. Gig Economy Coverage Framework

Develop portable benefits systems for independent contractors and part-time workers. Western states' elevated uninsured rates reflect changing labor markets that traditional employer-based coverage cannot address.

### 3. Administrative Simplification

Eliminate Medicaid work requirements and streamline enrollment processes. The Arkansas case study demonstrates that administrative barriers reverse coverage gains and disproportionately harm vulnerable populations.

### 4. Targeted Affordability Support

Expand premium subsidies beyond current income thresholds, as the middle-income analysis reveals that affordability challenges persist across income levels due to varying state costs and policies.

# Appendix

## A.1 References

-   Davis, K. (2007). *Uninsured in America: Problems and possible solutions*. *BMJ*, 334(7589), 346–348. <https://doi.org/10.1136/bmj.39091.493588.be>
-   KFF (2023). *Key Facts About the Uninsured Population*. <https://www.kff.org/uninsured/issue-brief/key-facts-about-the-uninsured-population/>
-   Sommers, B. D., et al. (2019). *Medicaid Work Requirements—Results from the First Year in Arkansas*. *New England Journal of Medicine*, 381(11), 1073–1082. <https://doi.org/10.1056/NEJMsr1901772>
-   Bailey, M. J., et al. (2017). *State Medicaid expansions and mortality, revisited: A cost-benefit analysis*. *American Journal of Health Economics*, 3(4), 392–421. <https://doi.org/10.1162/AJHE_a_00080>

<!-- -->

-   Collins, S. R., Gunja, M. Z., & Aboulafia, G. N. (2020). *U.S. health insurance coverage in 2020: A looming crisis in affordability*. *The Commonwealth Fund*.

<!-- -->

-   Sommers, B. D., Baicker, K., & Epstein, A. M. (2012). *Mortality and access to care among adults after state Medicaid expansions*. *New England Journal of Medicine*, 367(11), 1025–1034. <https://doi.org/10.1056/NEJMsa1202099>

## A.2 Session Info

<https://github.com/catherinamikhail/Programming-for-Economists-uninsured-US-population.git>
