---
title: "Visualizing the Uninsured: A Data Science Perspective on U.S. Health Coverage"
author: |
  - Eduardo Sáenz-Messía Laborda, 2861597
  - Catherina Mikhail, 2847118
  - Charlotte Craenen, 2865572
  - Hugo van As, 2848225
  - Indy Pleijter, 2810379
  - Julian Zeguers, 2772646
  - Wies Polderman, 2824987
  - Tutorial group 5, group 4 – J.F. Fitzgerald
date: "2025-06-24"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(repos = c(CRAN = "https://cran.rstudio.com/"))

packages <- c("rmarkdown", "tinytex", "rnaturalearth", "rnaturalearthdata", "maps", "tidyverse", "tidyr", "dplyr", "renv", "rnaturalearthhires")

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) {
  install.packages(packages[!installed], repos = "https://cran.rstudio.com/")
}

invisible(lapply(packages, library, character.only = TRUE))

if (!tinytex::is_tinytex()) { 
  tinytex::install_tinytex()
}

if (!require("remotes")) {
  install.packages("remotes")
}
if(!
   require("rnaturalearthhires")){
  remotes:: install_github("reponsci/rnaturalearthhires")
}
library(rnaturalearthhires)
```

**Part 1 – Identify a Social Problem**

## Describe the Social Problem

The substantial number of uninsured people in the United States represents a significant public health challenge with serious implications for individuals and the healthcare system. Current data shows that over 27 million Americans, approximately one in twelve people, lack health insurance coverage as of 2022 (KFF, 2023). This coverage gap creates considerable barriers to healthcare access and produces widespread effects on both individual and population health outcomes.

For uninsured individuals, the consequences are particularly severe. Research indicates that people without insurance are significantly more likely to postpone or forgo necessary medical care due to cost concerns, leading to poorer health outcomes, increased emergency department utilization for preventable conditions, and higher mortality rates from treatable diseases. The economic burden is equally substantial, as uninsured individuals face full payment responsibility for medical services, frequently resulting in medical debt, compromised credit status, and prolonged financial hardship (Davis, 2007).

The impact extends beyond individual consequences to affect the broader healthcare system. When uninsured patients cannot meet their payment obligations, healthcare institutions must absorb these costs as uncompensated care. This financial burden creates strain on healthcare providers and contributes to cost increases that ultimately affect insured populations through higher premiums and increased public spending. Davis (2007) emphasizes that the uninsured population includes "our neighbors, co-workers, and family members," demonstrating how this issue affects entire communities and has implications for society as a whole.

## Provide Background on the Problem

The Affordable Care Act (ACA), implemented in 2010, substantially expanded health insurance coverage through Medicaid expansion and subsidized insurance marketplaces. Nevertheless, millions of Americans continue to lack insurance coverage despite these policy reforms. According to the Kaiser Family Foundation, the majority of uninsured individuals today are low-income workers, with disproportionate representation among communities of color and residents of states that declined to expand Medicaid under the ACA (KFF, 2023).

A primary contributing factor to persistent uninsurance is the employment-based structure of the U.S. health insurance system. Many uninsured individuals are employed in positions that do not provide health benefits, including part-time, temporary, or service sector roles. These workers frequently earn incomes that exceed traditional Medicaid eligibility thresholds while remaining insufficient to afford private insurance premiums, creating what Davis (2007) characterizes as falling "through the cracks" of the existing system.

Geographic variation in coverage also contributes significantly to uninsurance patterns. States that implemented Medicaid expansion under the ACA demonstrate considerably lower uninsured rates compared to non-expansion states, resulting in unequal access to coverage nationwide. This policy divergence has produced particularly elevated uninsured rates in numerous Southern and certain Western states.

The continued prevalence of uninsurance reflects fundamental structural limitations within American healthcare policy, including affordability constraints, accessibility barriers, and coverage gaps that leave vulnerable populations without viable insurance options.

# Part 2 – Describe and Acquire Data

## Describe the Data sets

This analysis employs two primary datasets to examine health insurance coverage patterns and their relationship to economic conditions across U.S. states.

**U.S. Census Bureau Data** The principal dataset derives from the U.S. Census Bureau's American Community Survey (ACS) 1-Year Estimates, specifically the table "Selected Characteristics of the Uninsured in the United States." This dataset encompasses the period from 2010 to 2023, with 2020 excluded due to the unavailability of 1-year estimates for that year.

Eight smaller states and territories were excluded from the analysis: Delaware, District of Columbia, Hawaii, Puerto Rico, North Dakota, Rhode Island, Vermont, and Wyoming, as they are not consistently represented in the 1-year estimates. The 1-year estimates were selected over 5-year estimates because they provide more current data capable of capturing short-term variations resulting from economic fluctuations or policy modifications.

The ACS dataset provides state-level information on the civilian non-institutionalized population, including total population counts, numbers of uninsured individuals, and various demographic and socioeconomic characteristics. This analysis focused on total population figures and household income data (inflation-adjusted) to calculate uninsured rates and assess the economic context.

**Bureau of Economic Analysis Data** Economic data originates from the Bureau of Economic Analysis (BEA) table SAGDP1 – State Annual Gross Domestic Product Summary. This dataset provides real GDP figures in millions of chained 2017 dollars for each state from 1997 through 2024. The use of inflation-adjusted GDP facilitates consistent comparisons across time periods and between different states.

**Data Integration** These two datasets collectively provide a comprehensive perspective on how health insurance coverage has evolved across the United States over time and how coverage patterns may correlate with state-level economic conditions. This combination enables analysis of both temporal trends and interstate variation in uninsured rates.

## Import and Prepare the Data set

**First data set:** **"Selected Characteristics of The Uninsured in the U.S."**

Here we imported the first data set from the U.S. Census Bureau and named them "data set_uninsured_population\_*year"* from 2010-2023, excl 2020.

```{r import-data, include=FALSE}
# The first dataset is from the dataset "selected characteristics of the uninsured in the US from 2010-2023, exclduing 2020 from the US Census Bureau. These are the ACS one-year estimates.

#2010
ACSST1Y2010_S2702_2025_06_03T123607 <- read_csv("data insured population 2010-2023/ACSST1Y2010.S2702-2025-06-03T123607.csv")
dataset_uninsured_population_2010 = read_csv("data insured population 2010-2023/ACSST1Y2010.S2702-2025-06-03T123607.csv")

#2011
ACSST1Y2011_S2702_2025_06_03T123539 <- read_csv("data insured population 2010-2023/ACSST1Y2011.S2702-2025-06-03T123539.csv")
dataset_uninsured_population_2011 = read_csv("data insured population 2010-2023/ACSST1Y2011.S2702-2025-06-03T123539.csv")

#2012
ACSST1Y2012_S2702_2025_06_03T123441 <- read_csv("data insured population 2010-2023/ACSST1Y2012.S2702-2025-06-03T123441.csv")
dataset_uninsured_population_2012 = read_csv("data insured population 2010-2023/ACSST1Y2012.S2702-2025-06-03T123441.csv")

#2013
ACSST1Y2013_S2702_2025_06_03T123404 <- read_csv("data insured population 2010-2023/ACSST1Y2013.S2702-2025-06-03T123404.csv")
dataset_uninsured_population_2013 = read_csv("data insured population 2010-2023/ACSST1Y2013.S2702-2025-06-03T123404.csv")

#2014
ACSST1Y2014_S2702_2025_06_03T123336 <- read_csv("data insured population 2010-2023/ACSST1Y2014.S2702-2025-06-03T123336.csv")
dataset_uninsured_population_2014 = read_csv("data insured population 2010-2023/ACSST1Y2014.S2702-2025-06-03T123336.csv")

#2015
ACSST1Y2015_S2702_2025_06_03T123312 <- read_csv("data insured population 2010-2023/ACSST1Y2015.S2702-2025-06-03T123312.csv")
dataset_uninsured_population_2015 = read_csv("data insured population 2010-2023/ACSST1Y2015.S2702-2025-06-03T123312.csv")

#2016
ACSST1Y2016_S2702_2025_06_03T123232 <- read_csv("data insured population 2010-2023/ACSST1Y2016.S2702-2025-06-03T123232.csv")
dataset_uninsured_population_2016 = read_csv("data insured population 2010-2023/ACSST1Y2016.S2702-2025-06-03T123232.csv")

#2017
ACSST1Y2017_S2702_2025_06_03T123200 <- read_csv("data insured population 2010-2023/ACSST1Y2017.S2702-2025-06-03T123200.csv")
dataset_uninsured_population_2017 = read_csv("data insured population 2010-2023/ACSST1Y2017.S2702-2025-06-03T123200.csv")

#2018
ACSST1Y2018_S2702_2025_06_03T123128 <- read_csv("data insured population 2010-2023/ACSST1Y2018.S2702-2025-06-03T123128.csv")
dataset_uninsured_population_2018 = read_csv("data insured population 2010-2023/ACSST1Y2018.S2702-2025-06-03T123128.csv")

#2019
ACSST1Y2019_S2702_2025_06_03T123018 <- read_csv("data insured population 2010-2023/ACSST1Y2019.S2702-2025-06-03T123018.csv")
dataset_uninsured_population_2019 = read_csv("data insured population 2010-2023/ACSST1Y2019.S2702-2025-06-03T123018.csv")

#2021
ACSST1Y2021_S2702_2025_06_03T122952 <- read_csv("data insured population 2010-2023/ACSST1Y2021.S2702-2025-06-03T122952.csv")
dataset_uninsured_population_2021 = read_csv("data insured population 2010-2023/ACSST1Y2021.S2702-2025-06-03T122952.csv")

#2022
ACSST1Y2022_S2702_2025_06_03T122917 <- read_csv("data insured population 2010-2023/ACSST1Y2022.S2702-2025-06-03T122917.csv")
dataset_uninsured_population_2022 = read_csv("data insured population 2010-2023/ACSST1Y2022.S2702-2025-06-03T122917.csv")

#2023
ACSST1Y2023_S2702_2025_06_04T165512 <- read_csv("data insured population 2010-2023/ACSST1Y2023.S2702-2025-06-04T165512.csv")
dataset_uninsured_population_2023 = read_csv("data insured population 2010-2023/ACSST1Y2023.S2702-2025-06-04T165512.csv")

```

**Second Data set: "SAGDP1 State Annual Gross Domestic Product Summary", Statistic: "Real GDP (millions of chained 2017 dollars)"**

Here we imported the second data set from the Bureau of Economic Analysis of the annual GDP of the states from 2009-2023 and named it "table_GDP"

```{r import second dataset, figure.show="asis", include=F}
#The second dataset is from the Bureau of Economic Analysis and is the annual real GDP of the 44 states we include in our analysis from the year 2009-2023, excluding 2020
annual_gdp_Table_4_ <- read_csv("data GDP/annual gdp - Table (4).csv")

#renaming the table
table_GDP = annual_gdp_Table_4_
```

## Data cleaning: Data sets of Uninsured Population and The Creation of The New Variable: The Uninsured Share

To prepare the data for analysis, we removed margins of error from the original Census datasets, as they were unnecessary for comparing overall trends in health insurance coverage across states and years. A structural challenge involved smaller states and territories, since the Census Bureau's 1-year estimates no longer cover certain low-population areas after 2016. To ensure consistency, we excluded eight jurisdictions: Delaware, the District of Columbia, Hawaii, Puerto Rico, North Dakota, Rhode Island, Vermont, and Wyoming. This ensured our dataset included only states present in every applicable year from 2010 to 2023, excluding 2020 when no data was published due to COVID-19.

We streamlined the data structure by removing extensive metadata not essential to our analysis, including variables related to demographics, citizenship status, education, employment, and poverty ratios. We retained only total civilian non-institutionalized population and household income data (inflation-adjusted), which were directly relevant to our research questions. We also converted numeric values stored as character strings into actual numeric variables to enable mathematical operations.

Most importantly, we created an uninsured share variable by dividing total uninsured individuals by total population and multiplying by 100 for each state and year. This percentage enables easier comparisons across time and states regardless of population differences. After these steps, we combined the individual yearly datasets into a single comprehensive dataset covering 2010 to 2023, excluding 2020, enabling consistent analysis of long-term trends.

```{r datacleaning of the yearly datasets Uninsured Population 2010,2023, include=FALSE}
# This is me cleaning out the data insured population from 2010-2016 getting out the margin of error
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2010)))]

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2011)))]

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2012)))]

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2013)))]

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2014)))]

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2015)))]

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2016)))]

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2017)))]

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2018)))]

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2019)))]

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2021)))]

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2022)))]

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Margin of Error", colnames(dataset_uninsured_population_2023)))]

###############################################################################

#This is me cleaning out the columns of Delaware
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2010)))]


dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2011)))]


dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2012)))]


dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2013)))]


dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2014)))]


dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2015)))]


dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2016)))]


dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2017)))]


dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2018)))]


dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2019)))]


dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2021)))]


dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2022)))]


dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Delaware", colnames(dataset_uninsured_population_2023)))]


#This is me cleaning out the columns of District of Columbia
dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2011)))]


dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2012)))]


dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2013)))]


dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2014)))]


dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2015)))]


dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("District of Columbia", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)

#This is me cleaning out the columns of Hawaii
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Hawaii", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of North Dakota
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("North Dakota", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Puerto Rico
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Puerto Rico", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Rhode Island
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Rhode Island", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Vermont
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Vermont", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)
#This is me cleaning out the columns of Wyoming
dataset_uninsured_population_2010 = dataset_uninsured_population_2010[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2010)))]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 = dataset_uninsured_population_2011[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2011)))]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 = dataset_uninsured_population_2012[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2012)))]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 = dataset_uninsured_population_2013[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2013)))]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 = dataset_uninsured_population_2014[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2014)))]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 = dataset_uninsured_population_2015[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2015)))]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 = dataset_uninsured_population_2016[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2016)))]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 = dataset_uninsured_population_2017[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2017)))]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 = dataset_uninsured_population_2018[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2018)))]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 = dataset_uninsured_population_2019[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2019)))]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 = dataset_uninsured_population_2021[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2021)))]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 = dataset_uninsured_population_2022[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2022)))]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 = dataset_uninsured_population_2023[, which(!grepl("Wyoming", colnames(dataset_uninsured_population_2023)))]
View(dataset_uninsured_population_2023)

###############################################################################

#Now we are cleaning out the unnecessary metadata in the rows
#this is me cleaning out the rows with unnecessary metadata and keeping household income levels and the total population numbers
dataset_uninsured_population_2010 <- dataset_uninsured_population_2010[-c(2:104),]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 <- dataset_uninsured_population_2011[-c(2:104),]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 <- dataset_uninsured_population_2012[-c(2:104),]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 <- dataset_uninsured_population_2013[-c(2:104),]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 <- dataset_uninsured_population_2014[-c(2:104),]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 <- dataset_uninsured_population_2015[-c(2:104),]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 <- dataset_uninsured_population_2016[-c(2:104),]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 <- dataset_uninsured_population_2017[-c(2:104),]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 <- dataset_uninsured_population_2018[-c(2:104),]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 <- dataset_uninsured_population_2019[-c(2:104),]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 <- dataset_uninsured_population_2021[-c(2:104),]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 <- dataset_uninsured_population_2022[-c(2:104),]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 <- dataset_uninsured_population_2023[-c(2:104),]
View(dataset_uninsured_population_2023)

dataset_uninsured_population_2010 <- dataset_uninsured_population_2010[-c(10:17),]
View(dataset_uninsured_population_2010)

dataset_uninsured_population_2011 <- dataset_uninsured_population_2011[-c(10:17),]
View(dataset_uninsured_population_2011)

dataset_uninsured_population_2012 <- dataset_uninsured_population_2012[-c(10:17),]
View(dataset_uninsured_population_2012)

dataset_uninsured_population_2013 <- dataset_uninsured_population_2013[-c(10:17),]
View(dataset_uninsured_population_2013)

dataset_uninsured_population_2014 <- dataset_uninsured_population_2014[-c(10:17),]
View(dataset_uninsured_population_2014)

dataset_uninsured_population_2015 <- dataset_uninsured_population_2015[-c(10:17),]
View(dataset_uninsured_population_2015)

dataset_uninsured_population_2016 <- dataset_uninsured_population_2016[-c(10:17),]
View(dataset_uninsured_population_2016)

dataset_uninsured_population_2017 <- dataset_uninsured_population_2017[-c(10:17),]
View(dataset_uninsured_population_2017)

dataset_uninsured_population_2018 <- dataset_uninsured_population_2018[-c(10:17),]
View(dataset_uninsured_population_2018)

dataset_uninsured_population_2019 <- dataset_uninsured_population_2019[-c(10:17),]
View(dataset_uninsured_population_2019)

dataset_uninsured_population_2021 <- dataset_uninsured_population_2021[-c(10:17),]
View(dataset_uninsured_population_2021)

dataset_uninsured_population_2022 <- dataset_uninsured_population_2022[-c(10:17),]
View(dataset_uninsured_population_2022)

dataset_uninsured_population_2023 <- dataset_uninsured_population_2023[-c(10:17),]
View(dataset_uninsured_population_2023)

################################################################################

#As the values in these yearly datasets are stored as characters, we are now going to coerce them into numeric values so that we can create a new variable afterwards.
#The values in the datasets of the uninsured population are stored as characters. Here I am going to get rid of "" so I can store them as numeric.

# Keep the first column as-is
first_col <- dataset_uninsured_population_2010[[1]]

# Clean and convert the rest of the columns
# Keep the first column as-is
first_col <- dataset_uninsured_population_2010[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2010_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2010[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2010_cleaned <- cbind(first_col, dataset_uninsured_population_2010_numeric)

#rename the first column
colnames(dataset_uninsured_population_2010_cleaned)[1] <- colnames(dataset_uninsured_population_2010)[1]

View(dataset_uninsured_population_2010_cleaned)
class(dataset_uninsured_population_2010_cleaned)
class(dataset_uninsured_population_2010_cleaned[[2]])




# Keep the first column as-is
first_col <- dataset_uninsured_population_2011[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2011_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2011[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2011_cleaned <- cbind(first_col, dataset_uninsured_population_2011_numeric)

#rename the first column
colnames(dataset_uninsured_population_2011_cleaned)[1] <- colnames(dataset_uninsured_population_2011)[1]

View(dataset_uninsured_population_2011_cleaned)
class(dataset_uninsured_population_2011_cleaned)
class(dataset_uninsured_population_2011_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2012[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2012_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2012[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2012_cleaned <- cbind(first_col, dataset_uninsured_population_2012_numeric)

#rename the first column
colnames(dataset_uninsured_population_2012_cleaned)[1] <- colnames(dataset_uninsured_population_2012)[1]

View(dataset_uninsured_population_2012_cleaned)
class(dataset_uninsured_population_2012_cleaned)
class(dataset_uninsured_population_2012_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2013[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2013_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2013[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2013_cleaned <- cbind(first_col, dataset_uninsured_population_2013_numeric)

#rename the first column
colnames(dataset_uninsured_population_2013_cleaned)[1] <- colnames(dataset_uninsured_population_2013)[1]

View(dataset_uninsured_population_2013_cleaned)
class(dataset_uninsured_population_2013_cleaned)
class(dataset_uninsured_population_2013_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2014[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2014_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2014[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2014_cleaned <- cbind(first_col, dataset_uninsured_population_2014_numeric)

#rename the first column
colnames(dataset_uninsured_population_2014_cleaned)[1] <- colnames(dataset_uninsured_population_2014)[1]

View(dataset_uninsured_population_2014_cleaned)
class(dataset_uninsured_population_2014_cleaned)
class(dataset_uninsured_population_2014_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2015[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2015_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2015[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2015_cleaned <- cbind(first_col, dataset_uninsured_population_2015_numeric)

#rename the first column
colnames(dataset_uninsured_population_2015_cleaned)[1] <- colnames(dataset_uninsured_population_2015)[1]

View(dataset_uninsured_population_2015_cleaned)
class(dataset_uninsured_population_2015_cleaned)
class(dataset_uninsured_population_2015_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2016[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2016_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2016[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2016_cleaned <- cbind(first_col, dataset_uninsured_population_2016_numeric)

#rename the first column
colnames(dataset_uninsured_population_2016_cleaned)[1] <- colnames(dataset_uninsured_population_2016)[1]

View(dataset_uninsured_population_2016_cleaned)
class(dataset_uninsured_population_2016_cleaned)
class(dataset_uninsured_population_2016_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2017[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2017_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2017[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2017_cleaned <- cbind(first_col, dataset_uninsured_population_2017_numeric)

#rename the first column
colnames(dataset_uninsured_population_2017_cleaned)[1] <- colnames(dataset_uninsured_population_2017)[1]

View(dataset_uninsured_population_2017_cleaned)
class(dataset_uninsured_population_2017_cleaned)
class(dataset_uninsured_population_2017_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2018[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2018_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2018[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2018_cleaned <- cbind(first_col, dataset_uninsured_population_2018_numeric)

#rename the first column
colnames(dataset_uninsured_population_2018_cleaned)[1] <- colnames(dataset_uninsured_population_2018)[1]

View(dataset_uninsured_population_2018_cleaned)
class(dataset_uninsured_population_2018_cleaned)
class(dataset_uninsured_population_2018_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2019[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2019_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2019[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2019_cleaned <- cbind(first_col, dataset_uninsured_population_2019_numeric)

#rename the first column
colnames(dataset_uninsured_population_2019_cleaned)[1] <- colnames(dataset_uninsured_population_2019)[1]

View(dataset_uninsured_population_2019_cleaned)
class(dataset_uninsured_population_2019_cleaned)
class(dataset_uninsured_population_2019_cleaned[[2]])

# Keep the first column as-is
first_col <- dataset_uninsured_population_2021[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2021_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2021[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2021_cleaned <- cbind(first_col, dataset_uninsured_population_2021_numeric)

#rename the first column
colnames(dataset_uninsured_population_2021_cleaned)[1] <- colnames(dataset_uninsured_population_2021)[1]

View(dataset_uninsured_population_2021_cleaned)
class(dataset_uninsured_population_2021_cleaned)
class(dataset_uninsured_population_2021_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2022[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2022_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2022[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2022_cleaned <- cbind(first_col, dataset_uninsured_population_2022_numeric)

#rename the first column
colnames(dataset_uninsured_population_2022_cleaned)[1] <- colnames(dataset_uninsured_population_2022)[1]

View(dataset_uninsured_population_2022_cleaned)
class(dataset_uninsured_population_2022_cleaned)
class(dataset_uninsured_population_2022_cleaned[[2]])


# Keep the first column as-is
first_col <- dataset_uninsured_population_2023[[1]]

# Clean and convert the rest of the columns
dataset_uninsured_population_2023_numeric <- as.data.frame(
  lapply(dataset_uninsured_population_2023[ , -1], function(col) {
    cleaned <- gsub("[\"%,]", "", col)
    as.numeric(cleaned)
  }),
  stringsAsFactors = FALSE
)

#Combine back with the first column
dataset_uninsured_population_2023_cleaned <- cbind(first_col, dataset_uninsured_population_2023_numeric)

#rename the first column
colnames(dataset_uninsured_population_2023_cleaned)[1] <- colnames(dataset_uninsured_population_2023)[1]

View(dataset_uninsured_population_2023_cleaned)
class(dataset_uninsured_population_2023_cleaned)
class(dataset_uninsured_population_2023_cleaned[[2]])

################################################################################

#Now that the values in the yearly datasets are stored as numeric values but the first column as the first column contains the names of the states, we can now create are first new variable, which is the uninsured share:

#Here I am trying to compute and create a new variable, which is the share of uninsured population per state and per year

# Copy original data 2010
df_pop_2010 <- dataset_uninsured_population_2010_cleaned
col_names <- colnames(df_pop_2010)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2010[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2010[[total_col]][1]) && df_pop_2010[[total_col]][1] != 0) {
      df_pop_2010[[share_col]][1] <- df_pop_2010[[uninsured_col]][1] / df_pop_2010[[total_col]][1]
    } else {
      df_pop_2010[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2010 <- df_pop_2010[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2010), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2010[1, share_cols])

View(df_pop_2010)

#################################################################################

# Copy original data 2011
df_pop_2011 <- dataset_uninsured_population_2011_cleaned
col_names <- colnames(df_pop_2011)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2011[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2011[[total_col]][1]) && df_pop_2011[[total_col]][1] != 0) {
      df_pop_2011[[share_col]][1] <- df_pop_2011[[uninsured_col]][1] / df_pop_2011[[total_col]][1]
    } else {
      df_pop_2011[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2011 <- df_pop_2011[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2011), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2011[1, share_cols])

View(df_pop_2011)


###############################################################################


# Copy original data 2012
df_pop_2012 <- dataset_uninsured_population_2012_cleaned
col_names <- colnames(df_pop_2012)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2012[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2012[[total_col]][1]) && df_pop_2012[[total_col]][1] != 0) {
      df_pop_2012[[share_col]][1] <- df_pop_2012[[uninsured_col]][1] / df_pop_2012[[total_col]][1]
    } else {
      df_pop_2012[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2012 <- df_pop_2012[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2012), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2012[1, share_cols])


###############################################################################


# Copy original data 2013
df_pop_2013 <- dataset_uninsured_population_2013_cleaned
col_names <- colnames(df_pop_2013)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2013[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2013[[total_col]][1]) && df_pop_2013[[total_col]][1] != 0) {
      df_pop_2013[[share_col]][1] <- df_pop_2013[[uninsured_col]][1] / df_pop_2013[[total_col]][1]
    } else {
      df_pop_2013[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2013 <- df_pop_2013[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2013), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2013[1, share_cols])


################################################################################


# Copy original data 2014
df_pop_2014 <- dataset_uninsured_population_2014_cleaned
col_names <- colnames(df_pop_2014)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2014[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2014[[total_col]][1]) && df_pop_2014[[total_col]][1] != 0) {
      df_pop_2014[[share_col]][1] <- df_pop_2014[[uninsured_col]][1] / df_pop_2014[[total_col]][1]
    } else {
      df_pop_2014[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2014 <- df_pop_2014[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2014), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2014[1, share_cols])


################################################################################


# Copy original data 2015
df_pop_2015 <- dataset_uninsured_population_2015_cleaned
col_names <- colnames(df_pop_2015)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2015[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2015[[total_col]][1]) && df_pop_2015[[total_col]][1] != 0) {
      df_pop_2015[[share_col]][1] <- df_pop_2015[[uninsured_col]][1] / df_pop_2015[[total_col]][1]
    } else {
      df_pop_2015[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2015 <- df_pop_2015[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2015), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2015[1, share_cols])


###############################################################################


# Copy original data 2016
df_pop_2016 <- dataset_uninsured_population_2016_cleaned
col_names <- colnames(df_pop_2016)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2016[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2016[[total_col]][1]) && df_pop_2016[[total_col]][1] != 0) {
      df_pop_2016[[share_col]][1] <- df_pop_2016[[uninsured_col]][1] / df_pop_2016[[total_col]][1]
    } else {
      df_pop_2016[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2016 <- df_pop_2016[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2016), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2016[1, share_cols])

################################################################################

#something went wrong here with 2017 and 2018, so the function that worked is much more below as the column names in the years of 2017-2018 were named differently than the years before

# Copy original data 2017
df_pop_2017 <- dataset_uninsured_population_2017_cleaned
col_names <- colnames(df_pop_2017)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2017[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2017[[total_col]][1]) && df_pop_2017[[total_col]][1] != 0) {
      df_pop_2017[[share_col]][1] <- df_pop_2017[[uninsured_col]][1] / df_pop_2017[[total_col]][1]
    } else {
      df_pop_2017[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2017 <- df_pop_2017[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2017), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2017[1, share_cols])

# Copy original data 2018
df_pop_2018<- dataset_uninsured_population_2018_cleaned
col_names <- colnames(df_pop_2018)
new_col_order <- col_names  # to control column ordering

# Loop through all columns ending in "..Uninsured.Population..Estimate"
uninsured_cols <- grep("\\.\\.Uninsured\\.Population\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract the state name — everything before the first ".."
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Define the total population column and the share column
  total_col <- paste0(state, "..Total.Civilian.Noninstitutionalized.Population..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  if (total_col %in% col_names) {
    # Initialize new share column with NA
    df_pop_2018[[share_col]] <- NA
    
    # Calculate only for first row
    if (!is.na(df_pop_2018[[total_col]][1]) && df_pop_2018[[total_col]][1] != 0) {
      df_pop_2018[[share_col]][1] <- df_pop_2018[[uninsured_col]][1] / df_pop_2018[[total_col]][1]
    } else {
      df_pop_2018[[share_col]][1] <- NA
    }
    
    # Insert new column right after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
    
  } else {
    cat("⚠️ Total population column not found for:", state, "\n")
  }
}

# Reorder the columns
df_pop_2018 <- df_pop_2018[, new_col_order]

# Preview: Show share columns for the first row
share_cols <- grep("\\.\\.Uninsured.Share$", colnames(df_pop_2018), value = TRUE)
cat("\n✅ Created share columns:\n")
print(share_cols)
print(df_pop_2018[1, share_cols])

View(df_pop_2017)
View(df_pop_2016)
View(df_pop_2018)

#I have seen that it did not work for the years of 2017 and 2018 so I have to see what the names of the columns are
grep("Alabama", colnames(dataset_uninsured_population_2017_cleaned), value = TRUE)
grep("Alabama", colnames(dataset_uninsured_population_2018_cleaned), value = TRUE)

#Here I am starting again from 2017

# Make a working copy of your data 2017
df_pop_2017 <- dataset_uninsured_population_2017_cleaned
col_names <- colnames(df_pop_2017)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2017[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2017[[total_col]][1]
    uninsured_val <- df_pop_2017[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2017[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2017 <- df_pop_2017[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2017), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2017[1, share_cols])

#now it worked and I can work further on 2018-2023 except for 2020 as we do not have the data of that

# Make a working copy of your data 2018
df_pop_2018 <- dataset_uninsured_population_2018_cleaned
col_names <- colnames(df_pop_2018)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2018[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2018[[total_col]][1]
    uninsured_val <- df_pop_2018[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2018[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2018 <- df_pop_2018[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2018), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2018[1, share_cols])


################################################################################


# Make a working copy of your data 2019
df_pop_2019 <- dataset_uninsured_population_2019_cleaned
col_names <- colnames(df_pop_2019)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2019[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2019[[total_col]][1]
    uninsured_val <- df_pop_2019[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2019[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2019 <- df_pop_2019[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2019), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2019[1, share_cols])


################################################################################


# Make a working copy of your data 2021
df_pop_2021 <- dataset_uninsured_population_2021_cleaned
col_names <- colnames(df_pop_2021)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2021[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2021[[total_col]][1]
    uninsured_val <- df_pop_2021[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2021[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2021 <- df_pop_2021[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2021), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2021[1, share_cols])

###############################################################################


# Make a working copy of your data 2022
df_pop_2022 <- dataset_uninsured_population_2022_cleaned
col_names <- colnames(df_pop_2022)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2022[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2022[[total_col]][1]
    uninsured_val <- df_pop_2022[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2022[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2022 <- df_pop_2022[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2022), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2022[1, share_cols])


################################################################################

# Make a working copy of your data 2023
df_pop_2023 <- dataset_uninsured_population_2023_cleaned
col_names <- colnames(df_pop_2023)
new_col_order <- col_names

# Loop through all columns ending with "..Total.Uninsured..Estimate"
uninsured_cols <- grep("\\.\\.Total\\.Uninsured\\.\\.Estimate$", col_names, value = TRUE)

for (uninsured_col in uninsured_cols) {
  # Extract state (everything before first "..")
  state <- strsplit(uninsured_col, "\\.\\.")[[1]][1]
  
  # Construct matching total population column and new share column
  total_col <- paste0(state, "..Total..Estimate")
  share_col <- paste0(state, "..Uninsured.Share")
  
  cat("Processing:", state, "\n")
  
  # Check if the total population column exists
  if (total_col %in% col_names) {
    # Create new share column, default NA
    df_pop_2023[[share_col]] <- NA
    
    # Calculate share only for first row
    total_val <- df_pop_2023[[total_col]][1]
    uninsured_val <- df_pop_2023[[uninsured_col]][1]
    
    if (!is.na(total_val) && total_val != 0) {
      df_pop_2023[[share_col]][1] <- uninsured_val / total_val
    }
    
    # Insert share column after the uninsured column
    pos <- match(uninsured_col, new_col_order)
    new_col_order <- append(new_col_order, share_col, after = pos)
  } else {
    cat("⚠️ Total column not found for", state, "\n")
  }
}

# Reorder columns
df_pop_2023 <- df_pop_2023[, new_col_order]

# Show all created share columns
share_cols <- grep("\\.\\.Uninsured\\.Share$", names(df_pop_2023), value = TRUE)
cat("\n✅ Created columns:\n")
print(share_cols)
print(df_pop_2023[1, share_cols])

################################################################################

#Now we have the new variable in the seperate datasets and now we are going to merge these yearly datasets into 1 dataset

# Define years
years_early <- 2010:2016
years_late <- c(2017:2019, 2021:2023)

# Function to process one year of data
process_df <- function(df, year, is_early) {
  label_col <- names(df)[1]
  state_cols <- names(df)[!names(df) %in% c(label_col, "Unnamed: 0")]
  states <- unique(str_extract(state_cols, "^[^.]+"))
  
  # Set correct suffixes depending on year
  if (is_early) {
    pop_suffix <- "..Total.Civilian.Noninstitutionalized.Population..Estimate"
    unins_suffix <- "..Uninsured.Population..Estimate"
    share_suffix <- "..Uninsured.Share"
  } else {
    pop_suffix <- "..Total..Estimate"
    unins_suffix <- "..Total.Uninsured..Estimate"
    share_suffix <- "..Uninsured.Share"
  }
  
  # Extract for each state
  rows <- lapply(states, function(state) {
    pop_col <- paste0(state, pop_suffix)
    unins_col <- paste0(state, unins_suffix)
    share_col <- paste0(state, share_suffix)
    
    # Get only total population row
    total_row <- df %>%
      filter(str_trim(.data[[label_col]]) == "Total population")
    
    if (nrow(total_row) == 0 || !(pop_col %in% names(df))) return(NULL)
    
    tibble(
      state = state,
      year = year,
      total_population = as.numeric(total_row[[pop_col]]),
      total_uninsured = as.numeric(total_row[[unins_col]]),
      uninsured_share = as.numeric(total_row[[share_col]])
    )
  })
  
  bind_rows(rows)
}

# Process all early years
dfs_early <- mget(paste0("df_pop_", years_early))
panel_early <- map2_dfr(dfs_early, years_early, ~process_df(.x, .y, is_early = TRUE))

# Process all late years
dfs_late <- mget(paste0("df_pop_", years_late))
#panel_late <- map2_dfr(dfs_late, years_late, ~process_df(.x, .y, is_early = FALSE))

early_states <- bind_rows(
  dfs_early, .id = "year") %>%
  pivot_longer(
    -c(year, "Label (Grouping)"),
    names_to = "tmp",
    values_to = "value") %>%
  separate(tmp,
           into = c("state", "series", "measure"),
           sep = "\\.\\.",
           remove = TRUE)

late_states <- bind_rows(
  dfs_late, .id = "year") %>%
  pivot_longer(
    -c(year, "Label (Grouping)"),
    names_to = "tmp",
    values_to = "value") %>%
  separate(tmp,
           into = c("state", "series", "measure"),
           sep = "\\.\\.",
           remove = TRUE)
# View

all_states <- rbind(late_states, early_states)

head(all_states)
###############################################################################

#cleaning out df_pop from the column of years to only remain the year numbers

unique(all_states$year)

all_states$year <- str_remove(all_states$year, "df_pop_")

#Now I am coercing the string into numeric so plotting becomes easier

all_states$year <- as.numeric(all_states$year)
str(all_states)

################################################################################
#Here I am multiplying the uninsured share by 100, such that they are representing percentages
all_states_100 = all_states %>%
  mutate(value = 
           ifelse(grepl("Uninsured.Share",
                        series), value*100, value))

#This is the dataset for the Uninsured Population we can use for the visualizations
#renaming a column to make it easier "Label (Grouping)" to "income_group"

library(dplyr)
all_states_100 = all_states_100 %>%
  rename(income_group = `Label (Grouping)`)
```

## Data cleaning: Annual GDP Table and The Creation of The New Variable: Annual Real GDP Growth per year and state

For the economic data from the Bureau of Economic Analysis, we used the SAGDP1 – State Annual Gross Domestic Product Summary table, which reports real GDP per state in millions of chained 2017 dollars. We cleaned this dataset to align with our health insurance data by removing columns for years outside our scope (1997 through 2008, plus 2020 and 2024) and filtering out states not included in our main dataset. We excluded Delaware, District of Columbia, Hawaii, North Dakota, Puerto Rico, Vermont, Wyoming, and Rhode Island, as well as regional aggregates such as New England, Mideast, Great Lakes, Plains, Southeast, Southwest, Rocky Mountain, and Far West. We converted the stored values to numeric format and renamed the cleaned dataset `table_GDP`.

From this cleaned table, we created a new variable: annual real GDP growth, calculated for each state and year by measuring the year-over-year percentage change in real GDP. This variable captures how rapidly each state's economy was expanding or contracting over time. Including GDP growth adds important economic context to our analysis, allowing us to explore whether changes in economic performance relate to shifts in health insurance coverage. We can examine whether periods of economic expansion correspond with improvements in coverage rates, providing a more comprehensive understanding of how economic conditions may influence healthcare access.

```{r include=FALSE}
#Here I am making a new variable which is the GDP growth rate of the states

#renaming the table
table_GDP = annual_gdp_Table_4_

#cleaning out the table of the years 1997:2008, 2020 and 2024 and removing the states/regions that are not included in our main data set

table_GDP = annual_gdp_Table_4_[, !(names(annual_gdp_Table_4_) %in% c("1997", "1998","1999","2000", "2001","2002","2003","2004","2005","2006","2007","2008","2020","2024"))]

table_GDP = table_GDP[-c(9,10,13,36,41,47,52,54,55,56,57,58,59,60),]
table_GDP = table_GDP[-c(1),]
table_GDP = table_GDP[-c(45),]

#class coercion of double to numeric of all columns but the first column
table_GDP[, -1] = lapply(table_GDP[, -1], as.numeric)

#checking the class
sapply(table_GDP, class)

#yes, all the numbers are now numeric instead of the first column with the names of the states


# Pivot longer
gdp_long <- table_GDP %>%
  pivot_longer(
    cols = -GeoName,
    names_to = "year",
    values_to = "gdp"
  ) %>%
  mutate(year = as.integer(year))

View(table_GDP)
View(gdp_long)

# Calculate GDP growth rate
gdp_growth <- gdp_long %>%
  group_by(GeoName) %>%
  arrange(year) %>%
  mutate(growth_rate = (gdp - lag(gdp)) / lag(gdp) * 100)

View(gdp_growth)

#delete rows from 2009
gdp_growth = gdp_growth[-c(1:45), ]
```

# Part 3 – Visualize and Analyze the Data

## Create Initial Visualizations

**Spatial Variation Visualization, U.S. Share of The Uninsured Population in 2023**

To examine the spatial distribution of uninsured populations across the United States, we mapped the geographic variation using American Community Survey (ACS) one-year estimates for available states. Initially, we created a subset of the comprehensive all_states_100 data set, focusing specifically on 2023 data to ensure temporal consistency. Subsequently, we generated separate visualizations for Alaska due to its substantial geographic scale, which would otherwise compromise the readability and detail of the continental United States map when displayed at the same resolution.

```{r visualize spatial, echo = FALSE, fig.cap = "", out.width = "80%"}
#Here I am making a sub dataframe for the spatial visualization of 2023 of all the states with available data

spatial_df = all_states %>%
  filter(year == 2023, series == "Uninsured.Share") %>%
  select(state, value)

#deleting unnecessary rows
spatial_df = spatial_df[-c(45:396),]

#Here I am trying to plot the spatial subset

#getting the US map
us_map = map_data("state")

#clean up states names in both datasets
us_map$region = tolower(us_map$region)
spatial_df$state = tolower(spatial_df$state)

#join the map data with the subset data
us_map = left_join(us_map, spatial_df, by = c("region" = "state"))

#plot the map
ggplot(us_map, aes(x = long, y = lat, group = group, fill = value)) + 
  geom_polygon (color = "black") +
  scale_fill_gradient(low = "lightblue", high = "red", na.value = "white")+
  ggtitle("Share of Uninsured Population in the US, 2023")+
  labs(fill = "(%) Share of Uninsured Population") + 
  theme_void() 

```

```{r alaska, echo = FALSE, fig.cap = ""}

#I still need the state Alaska but the package maps does not include the state Alaska

# I am trying it here again including Alaska with another package

library(rnaturalearth)
library(ggplot2)
library(dplyr)

#get US shapefile
us_states = ne_states(country = "United States of America", returnclass = "sf")

#convert state names into lowercase letters such that they match
us_states$name = tolower(us_states$name)

alaska = us_states[us_states$name == "alaska",]

#merge data with shapefile
alaska_data = alaska %>%
  left_join(spatial_df, by = c("name" = "state"))

#plot the map with the data
ggplot(alaska_data) + 
  geom_sf(aes(fill = value)) + 
  theme_void() + 
  scale_fill_gradient( low = "lightblue", high = "red", na.value = "white")+
  theme(legend.position = "none")
```

```{r visualization temporal, echo = FALSE, fig.cap = ""}
#Here I am making a sub dataframe for the temporal visualization with the GDP growth rate of Texas and another subset of the share of uninsured population in the US over the years

#subset Texas share of uninsured population
texas_share = all_states_100 %>%
  filter(state == "Texas",
         year %in% 2010:2023,
         series == "Uninsured.Share")

#cleaning out the NA rows of subset Texas
texas_share = texas_share[-c(2:9),]
texas_share = texas_share[-c(3:10),]
texas_share = texas_share[-c(4:11),]
texas_share = texas_share[-c(5:12),]
texas_share = texas_share[-c(6:13),]
texas_share = texas_share[-c(7:14),]
texas_share = texas_share[-c(8:15),]
texas_share = texas_share[-c(9:16),]
texas_share = texas_share[-c(10:17),]
texas_share = texas_share[-c(11:18),]
texas_share = texas_share[-c(12:19),]
texas_share = texas_share[-c(13:20),]
texas_share = texas_share[-c(14:21),]


#subset GDP growth of texas of the years 2010-2023 except for 2020
library(dplyr)
texas_gdp = gdp_growth %>%
  filter(GeoName == "Texas",
         year %in% 2010:2023) %>%
  select(year, growth_rate)

#Here I will plot the temporal visualization of Texas over the years 2010-2023 except for year 2020

#the subsets I will use are
View(texas_share)
View(texas_gdp)

library(ggplot2)
library(dplyr)

#Add a type column to distinguish between GDP growth rate and Uninsured Share
texas_gdp$type = "GDP growth rate"
texas_share$type = "Uninsured Share"

#rename the columns to match
texas_share = texas_share %>%
  rename(gdp = value)

#combine the 2 dataframes
combined_data_texas = bind_rows(texas_gdp %>%
                                  select(year, gdp = growth_rate, type),
                                texas_share %>%
                                  select(year, gdp, type))

################################################################################

#plotting the lines from the 2 dataframes

ggplot() + 
  geom_line(data = texas_gdp, aes(x = year, y = growth_rate, color = "GDP Growth Rate"), size =1) +
  geom_line(data = texas_share, aes(x = year, y = gdp, color = "Uninsured Share"), size = 1) + 
  labs(title = "GDP growth rate and Uninsured Share in Texas 2010 - 2023, except 2020", x = "Year", y = "Percentage Points", color = "Type") +
  scale_y_continuous(breaks = seq(-2.5,25, by = 2.5), limits = c(-2.5,25)) + 
  scale_x_continuous(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2021,2022,2023)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

**Temporal Variation Visualization, Texas's Share of Uninsured Population and Texas's GDP growth rate over time**

Texas emerged as a notable outlier in our spatial variation analysis, prompting us to conduct a more comprehensive temporal examination of the state's uninsured population dynamics. Consequently, we analyzed Texas's proportion of uninsured residents across a thirteen-year period from 2010 to 2023, excluding 2020 data due to potential pandemic-related anomalies. Furthermore, we sought to investigate whether economic growth served as a contributing factor in explaining the observed patterns in insurance coverage rates throughout this time frame.

**Sub-Group Variation Visualization, The Uninsured Rates of the Middle-Income Group in 2017 in the U.S.**

Building upon findings from the Kaiser Family Foundation (2024), which established that uninsured individuals in the United States are disproportionately represented among low-income populations, we sought to examine the insurance coverage status of middle-income groups who often find themselves in a precarious position. Specifically, this demographic typically earns insufficient income to afford private insurance yet exceeds the eligibility thresholds for Medicaid assistance. Moreover, we selected 2017 as our focal year, representing the first full year following President Trump's inauguration in 2016, thereby capturing potential early policy implications on this vulnerable middle-income segment.

```{r visualization subgroup, echo = FALSE}
#create a subset of data 2017 for total unsinred share in the states

library(dplyr)

subgroup_total = all_states_100 %>%
  filter(year == 2017,
         series == "Uninsured.Share",
         income_group == "Total civilian noninstitutionalized population")
# create a subset data of 2017 and middle income
middle_income = all_states_100[ all_states_100$income_group == ("$25,000 to $49,999"),]
middle_income = middle_income[middle_income$series == "Total.Uninsured",]
#subset that will be used is middle_income of 2017
#calculate quartiles
quartiles = quantile(middle_income$value, probs = c(0, 0.25, 0.5, 0.75, 1))

middle_income_1 = middle_income[middle_income$value >= quartiles[1] & middle_income$value <= quartiles[2],]
middle_income_2 = middle_income[middle_income$value > quartiles[2] & middle_income$value <= quartiles [3],]
middle_income_3 = middle_income[middle_income$value > quartiles[3] & middle_income$value <= quartiles [4], ]
middle_income_4 = middle_income[middle_income$value > quartiles[4] & middle_income$value <= quartiles [5],]

#merge subsets into 1 dataframe
merged_subgroup = rbind(middle_income_1, middle_income_2, middle_income_3, middle_income_4)
 #add a column to indicate the quartile
merged_subgroup$quartile = factor(rep(c("Bottom", "Second","Third", "Top"), times = c(nrow(middle_income_1), nrow (middle_income_2), nrow(middle_income_3), nrow(middle_income_4))))

boxplot_data = merge(subgroup_total,
                     merged_subgroup, by = "state")

#print the quartiles
quartiles = quantile(merged_subgroup$value, probs = c(0.25, 0.5, 0.75))
print(paste("First quartile (Q1):", quartiles[1]))
print(paste("Second quartile (Q2):", quartiles[2]))
print(paste("Third quartile (Q3):", quartiles[3]))

#plot the 4 boxplots in 1 plot

library(ggplot2)

ggplot(boxplot_data, aes(x = factor(quartile), y = value.x)) + 
  geom_boxplot()+
  labs(x = "Quartiles in Proportion State Population, making $25k to $49,999", y = "Percentages Point Total Uninsured U.S.", title = "Uninsured Share among Middle-Income in the US, 2017") + 
  theme_classic() + scale_y_continuous(breaks = seq(0,20, by = 5), limits = c(0,20))
```

**Event Analysis Visualization, The Impact of The Implementation of Work Requirements for Medicaid in Arkansas, 2018**

In 2018, Arkansas became the first state to implement work requirements for Medicaid recipients as part of its Arkansas Works program, mandating that able-bodied adults aged 19-49 demonstrate at least 80 hours of monthly employment, education, or training activities to maintain their Medicaid eligibility. This policy initiative was designed to encourage workforce participation and reduce dependency on government assistance programs. To evaluate the causal impact of this intervention, we employed a comparative analysis framework where Arkansas serves as the treatment group, while the remaining 43 states with available data constitute the control group. Subsequently, we examined whether the implementation of work requirements produced statistically significant changes in Arkansas's uninsured rates relative to the weighted average change observed across the control states, thereby isolating the policy's specific effects from broader national trends in insurance coverage.

```{r visualization event, echo = FALSE, fig.cap = ""}
#Here I am making a sub dataframe for the event analysis of the implementation of stricter conditions for Medicaid in 2018 in Arkansas

#sub dataframe Arkansas share uninsured for all the years
arkansas_event =  all_states %>%
  filter(state == "Arkansas", series == "Uninsured.Share") %>%
  select(state,year, value)

#deleting rows
arkansas_event = arkansas_event[-c(2:9),]
arkansas_event = arkansas_event[-c(3:10),]
arkansas_event = arkansas_event[-c(4:11),]
arkansas_event = arkansas_event[-c(5:12),]
arkansas_event = arkansas_event[-c(6:13),]
arkansas_event = arkansas_event[-c(7:14),]
arkansas_event = arkansas_event[-c(8:15),]
arkansas_event = arkansas_event[-c(9:16),]
arkansas_event = arkansas_event[-c(10:17),]
arkansas_event = arkansas_event[-c(11:18),]
arkansas_event = arkansas_event[-c(12:19),]
arkansas_event = arkansas_event[-c(13:20),]
arkansas_event = arkansas_event[-c(14:21),]

View(arkansas_event)

#Here I compute the weighted average of the other 43 states excluding Arkansas
totals = all_states_100 %>%
  filter((str_detect(income_group, "Total civilian noninstitutionalized population")|str_detect(income_group, "Total population")) & state!= "Arkansas") %>%
  select(year, state,total = value, series)

uninsured_share = all_states_100 %>%
  filter(series == "Uninsured.Share" & state!="Arkansas") %>%
  select(year, state, uninsured_share = value)

weighted_avg_2 = totals %>%
  inner_join(uninsured_share, by = c("year", "state"))%>%
  group_by(year)%>%
  summarise(weighted_avg = sum(uninsured_share * total, na.rm = TRUE)/ sum(total, na.rm = TRUE))

#plotting the event
library(ggplot2)

#adding a vertical line at the specific year

event_year = 2018
#The subsets we must use are weighted_avg_2 and arkansas_event

ggplot() + geom_line(data = weighted_avg_2, aes(x = year, y = weighted_avg, color = "Weighted Average"), size = 1) + 
  geom_line(data = arkansas_event, 
            aes(x = year, y = value, color = "Arkansas Share Uninsured"),
            size = 1
            ) + 
  labs(x = "Year", y = "Percentages Point of the Uninsured Population", title = "AR Medicaid Work Requirements, 2018") + 
  scale_color_manual(values = c("Weighted Average" = "darkblue", "Arkansas Share Uninsured" = "darkred")) +
  theme_classic() + geom_vline (xintercept = event_year, color = "black") + annotate("text", x = event_year + 0.3 , y = max(c(weighted_avg_2$weighted_avg, arkansas_event$value)),
                                                                                     label = "Implementation \nRequirements", hjust = 0, size =3, fontface = "bold") +
  scale_x_continuous(breaks = c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2021,2022,2023)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Identify Trends and Patterns

**Spatial Variation Visualization, U.S. Share of The Uninsured Population in 2023**

The spatial distribution of uninsured rates across the U.S. in 2023 reveals a pronounced regional divide. Southern states, particularly Texas and Florida, report the highest uninsured shares. In contrast, Northeastern and Upper Midwestern states demonstrate substantially lower rates, likely attributable to more expansive Medicaid programs, higher employer-based coverage, and more generous state-level health initiatives (Bailey et al., 2017).

Western states like Nevada and Arizona present a more complex pattern, exhibiting elevated uninsured rates despite relatively progressive policy environments. This may reflect distinct labor market characteristics, including higher prevalence of gig work, part-time employment, and self-employment, which are associated with reduced access to employer-sponsored insurance (Collins et al., 2020).

These findings support existing literature demonstrating that health coverage in the U.S. is not solely determined by state policy, but is fundamentally linked to employment structures, income levels, and demographic characteristics (Davis, 2007; Sommers et al., 2012). Consequently, even among states with similar political orientations, uninsured rates can vary significantly based on labor market and population dynamics.

**Temporal Variation Visualization, Texas's Share of Uninsured Population and Texas's GDP growth rate over time**

The line graph illustrates the relationship between GDP growth rate and uninsured share in Texas from 2010 to 2023, excluding 2020. We observe a clear downward trend in the uninsured share, declining from approximately 23% to 17% over this period. However, despite economic fluctuations, particularly a GDP growth decline around 2016 and a spike in between 2017 and 2018, the uninsured rate remained relatively high and stable, especially after 2016.

This pattern suggests that economic growth alone does not ensure increased health insurance coverage. Even during years with strong GDP growth, Texas maintained one of the highest uninsured rates nationally. A likely explanation is Texas's decision not to expand Medicaid under the ACA, leaving many low-income adults without affordable coverage options (KFF, 2023; Davis, 2007). Policy choices, rather than solely economic performance, fundamentally shape insurance access.

**Sub-Group Variation Visualization, The Uninsured Rates of the Middle-Income Group in 2017 in the U.S.**

The box plot displays total uninsured share per U.S. state in 2023, grouped by quartiles of middle-income earners (those earning \$25,000–\$49,999). The y-axis reflects overall uninsured rates, not exclusively middle-income individuals. The second and third quartiles demonstrate nearly identical median uninsured rates, suggesting that higher proportions of middle-income residents do not clearly relate to higher or lower uninsured shares.

This supports the premise that income alone does not determine insurance coverage. Other factors including cost of living, Medicaid eligibility requirements, and state-level policy differences play substantial roles. Two states may have similar income distributions yet vastly different uninsured rates due to variations in insurance affordability, healthcare access, or Medicaid expansion status (Davis, 2007; KFF, 2023).

**Event Analysis Visualization, The Impact of The Implementation of Work Requirements for Medicaid in Arkansas, 2018**

The graph shows uninsured share in Arkansas compared to the national weighted average from 2010 to 2023, excluding Arkansas. Arkansas experienced a sharp decline in uninsured rates after expanding Medicaid under the Affordable Care Act, reaching a low around 2016–2017. However, after implementing work requirements for Medicaid in 2018, this trend reversed: Arkansas's uninsured rate increased while the national average remained low and stable.

This suggests the policy negatively affected coverage. Although intended to encourage employment, research indicates many people lost coverage not due to unwillingness to work, but because of administrative barriers and confusion about reporting requirements (Sommers et al., 2019). Consequently, thousands of eligible individuals were removed from Medicaid, increasing the uninsured share. The work requirements likely influenced uninsured percentages by creating new obstacles to maintaining insurance coverage, particularly for low-income adults.

# Part 4 – Communicate Findings

## Summarize Key Insights

To summarize our key insights, the spatial variation analysis demonstrated that regional policy choices and labor market structures significantly influence uninsured rates beyond political ideology alone. Western states like Nevada and Arizona exhibited elevated uninsured rates despite progressive policy environments, reflecting distinct employment patterns including higher prevalence of gig work and self-employment that reduce access to employer-sponsored insurance. The temporal analysis revealed that economic growth does not automatically translate into improved health coverage, particularly in non-expansion states like Texas, where strong GDP growth coincided with persistently high uninsured rates due to the absence of Medicaid expansion.

The subgroup analysis provided evidence that income distribution alone does not fully explain variation in uninsured rates across states. States with similar proportions of middle-income earners demonstrated vastly different uninsured rates, indicating that factors such as cost of living, state-specific policies, and insurance market accessibility play substantial roles in coverage outcomes. Finally, the event analysis of Arkansas's Medicaid work requirements illustrated how administrative barriers can reverse coverage gains, with the policy leading to increased uninsured rates as eligible individuals lost coverage due to reporting difficulties and employment instability rather than unwillingness to work. These findings collectively underscore that achieving universal coverage requires comprehensive policy approaches that address structural employment changes, administrative complexity, and regional economic variations rather than relying solely on economic growth or income-based solutions.

## Propose Solutions or Policy Recommendations

## Key Policy Solutions

Based on the analysis findings, four targeted policy interventions are recommended to address persistent uninsurance across the United States.

Enhanced federal funding should be provided to encourage remaining non-expansion states like Texas and Florida to adopt Medicaid expansion. The temporal analysis of Texas demonstrates that economic growth alone cannot close coverage gaps without accompanying policy action. Despite periods of strong GDP growth, Texas maintained one of the highest uninsured rates nationally, suggesting that macroeconomic performance is insufficient to guarantee insurance access. Federal incentives could include increased matching funds or temporary enhanced federal medical assistance percentages to make expansion financially attractive for holdout states. This approach would address the fundamental policy gap that leaves millions of low-income adults without affordable coverage options.

Portable benefits systems should be developed for independent contractors and part-time workers to address the changing nature of employment. Western states' elevated uninsured rates reflect evolving labor markets that traditional employer-based coverage cannot adequately address, particularly the rise of gig work, part-time employment, and self-employment. These workers often fall into coverage gaps, earning too much for traditional Medicaid but lacking access to employer-sponsored insurance. Policy solutions could include portable benefit accounts that workers maintain across jobs, expanded association health plans for freelancers, or modified marketplace subsidies specifically designed for non-traditional employment arrangements.

Medicaid work requirements should be eliminated and enrollment processes streamlined to remove administrative barriers that impede coverage access. The Arkansas case study provides compelling evidence that work requirements reverse coverage gains and disproportionately harm vulnerable populations. Research indicates that many people lost coverage not due to unwillingness to work, but because of administrative hurdles and confusion about reporting requirements. Simplifying enrollment procedures, reducing paperwork burdens, and eliminating punitive work requirements would help maintain coverage continuity and ensure that eligible individuals can access and retain benefits without navigating unnecessarily complex bureaucratic processes.

Premium subsidies should be expanded beyond current income thresholds to address affordability challenges that persist across income levels. The middle-income analysis reveals that affordability problems are not confined to the lowest income brackets, as state-level variations in costs and policies create different financial pressures across regions. Current subsidy structures may not adequately account for geographic cost variations or the reality that middle-income families in high-cost states face significant premium burdens. Expanding eligibility for premium tax credits and cost-sharing reductions, potentially through sliding scale adjustments based on regional cost indices, would help address coverage gaps among moderate-income populations who currently find insurance unaffordable despite not qualifying for existing assistance programs.

# Appendix

## A.1 References

-   Davis, K. (2007). *Uninsured in America: Problems and possible solutions*. *BMJ*, 334(7589), 346–348. <https://doi.org/10.1136/bmj.39091.493588.be>
-   KFF (2023). *Key Facts About the Uninsured Population*. <https://www.kff.org/uninsured/issue-brief/key-facts-about-the-uninsured-population/>
-   Sommers, B. D., et al. (2019). *Medicaid Work Requirements—Results from the First Year in Arkansas*. *New England Journal of Medicine*, 381(11), 1073–1082. <https://doi.org/10.1056/NEJMsr1901772>
-   Bailey, M. J., et al. (2017). *State Medicaid expansions and mortality, revisited: A cost-benefit analysis*. *American Journal of Health Economics*, 3(4), 392–421. <https://doi.org/10.1162/AJHE_a_00080>
-   Collins, S. R., Gunja, M. Z., & Aboulafia, G. N. (2020). *U.S. health insurance coverage in 2020: A looming crisis in affordability*. *The Commonwealth Fund*.
-   Sommers, B. D., Baicker, K., & Epstein, A. M. (2012). *Mortality and access to care among adults after state Medicaid expansions*. *New England Journal of Medicine*, 367(11), 1025–1034. <https://doi.org/10.1056/NEJMsa1202099>

## A.2 Session Info

<https://github.com/catherinamikhail/Programming-for-Economists-uninsured-US-population.git>
